---
title: "The SOM shortlist"
subtitle: "The variants that earned a second look, every quality measure, and the maps themselves: read from the cache of som_next_steps.Rmd, nothing trained here"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: hide
    mathjax: null
params:
  seeds: 5
  rlen: 10
  size: 10
  n_short: 5
---

```{css, echo = FALSE}
div.main-container { max-width: 1240px; }
body { font-size: 16px; line-height: 1.6; color: #0b0b0b; }
.box { border-left: 4px solid #2a78d6; background: #f3f7fd; padding: 10px 18px; margin: 16px 0 22px; }
.box p:last-child, .box ul:last-child { margin-bottom: 0; }
.box table { font-size: 13.5px; width: auto; } .box table td, .box table th { padding: 4px 9px; vertical-align: top; white-space: nowrap; }
.note { color: #52514e; font-size: 14.5px; }
table.dataTable { font-size: 13.5px; }
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center", out.width = "100%", comment = "#>")
library(plotly); library(DT); library(htmltools)
```

```{r code, results = "hide"}
# the cache of som_next_steps.Rmd: the same settings give the same key, so nothing is trained here
if (!exists("params")) { header <- rmarkdown::yaml_front_matter("som_shortlist.Rmd")$params; params <- lapply(header, function(x) if (is.list(x)) x$value else x) }
p <- params
Sys.setenv(PKG_LIBS = "-lquadmath -lmpfr -lgmp")
Rcpp::sourceCpp("next_steps/ns_core.cpp", cacheDir = "next_steps/.cpp_cache")
for (f in c("ns_data.R", "ns_maps.R", "ns_quality.R", "ns_bench.R", "ns_report.R")) source(file.path("next_steps/R", f))
datasets <- load_datasets()
SIZES <- c(5, 7, 10, 14); seeds <- seq_len(p$seeds)
res <- compute_all(datasets, VARIANTS, SIZES, seeds, p$rlen, cores = 1)   # cached; cores only matter on a miss
bench <- res$bench; fits <- res$fits; det <- res$detection
SHORT <- c(cbf = "CBF", pd = "Partial discharges", gestures = "Gestures", gunpoint = "GunPoint")[names(datasets)]
REF <- p$size
```

```{r select, include = FALSE}
# ----------------------------------------------------------------- the measures, each read with a fair judge
MEAS <- list(
  list(id = "EV",     m = "EV",     j = "euclid", higher = TRUE,  label = "EV, explained variance (Euclidean)"),
  list(id = "TE",     m = "TE",     j = "own",    higher = FALSE, label = "TE, topographic error (own distance)"),
  list(id = "KL",     m = "KL",     j = "dtw",    higher = FALSE, label = "KL, Kaski-Lagus error (DTW)"),
  list(id = "TP",     m = "TP",     j = "own",    higher = FALSE, label = "|TP|, topographic product (own distance)"),
  list(id = "purity", m = "purity", j = "own",    higher = TRUE,  label = "Purity (own distance)"),
  list(id = "ARI",    m = "ARI",    j = "own",    higher = TRUE,  label = "ARI (own distance)"),
  list(id = "NMI",    m = "NMI",    j = "own",    higher = TRUE,  label = "NMI (own distance)"),
  list(id = "trust",  m = "trust",  j = "euclid", higher = TRUE,  label = "Trustworthiness"),
  list(id = "cont",   m = "cont",   j = "euclid", higher = TRUE,  label = "Continuity"),
  list(id = "used",   m = "used",   j = "own",    higher = TRUE,  label = "Share of units used"))
names(MEAS) <- sapply(MEAS, `[[`, "id")
val <- function(ds, v, size, me) { b <- bench[bench$dataset == ds & bench$variant == v & bench$size == size & bench$judge == MEAS[[me]]$j, ]
  b <- b[order(b$seed), ]; x <- b[[MEAS[[me]]$m]]; if (me == "TP") abs(x) else x }
mean_of <- function(ds, v, size, me) { x <- val(ds, v, size, me); if (length(x)) mean(x, na.rm = TRUE) else NA }
det_mean <- function(v) { x <- det$AUROC[det$variant == v]; if (length(x)) mean(x) else NA }

# ----------------------------------------------------------------- the shortlist: a composite rank at REF x REF
# For every data set, rank the seventeen variants (1 = best) on each measure that judges the map's use
# (ARI, NMI, TE, KL, trustworthiness, continuity) and average the six ranks; rank them once more on the
# detection AUROC of the partial discharges; the mean of the four blocks is the composite. The n_short
# lowest join the shortlist, with every variant that is the best of the seventeen on ARI for some data set
# (the classes are the point of the maps), and the two controls.
RANK_ON <- c("ARI", "NMI", "TE", "KL", "trust", "cont")
vids <- names(VARIANTS)
rank_of <- function(k, me) { vals <- sapply(vids, mean_of, ds = k, size = REF, me = me); rank(if (MEAS[[me]]$higher) -vals else vals, na.last = "keep") }
blocks <- sapply(names(datasets), function(k) rowMeans(sapply(RANK_ON, rank_of, k = k), na.rm = TRUE))
det_rank <- rank(-sapply(vids, det_mean), na.last = "keep")
rank_tab <- data.frame(variant = vids, label = sapply(vids, function(v) VARIANTS[[v]]$label), blocks, detection = det_rank,
                       mean_rank = rowMeans(cbind(blocks, det_rank), na.rm = TRUE),
                       ari = sapply(vids, function(v) mean(sapply(names(datasets), mean_of, v = v, size = REF, me = "ARI"))),
                       te = sapply(vids, function(v) mean(sapply(names(datasets), mean_of, v = v, size = REF, me = "TE"))),
                       auroc = sapply(vids, det_mean), stringsAsFactors = FALSE)
rank_tab <- rank_tab[order(rank_tab$mean_rank), ]
ari_best <- sapply(names(datasets), function(k) names(which.min(rank_of(k, "ARI"))))
shortlist <- unique(c(head(rank_tab$variant[!rank_tab$variant %in% c("euclid", "dtw")], p$n_short), ari_best, "euclid", "dtw"))
shortlist <- shortlist[order(match(shortlist, rank_tab$variant))]
SL_COL <- setNames(c("#2a78d6", "#eb6834", "#1baf7a", "#eda100", "#e87ba4", "#008300", "#4a3aa7", "#e34948")[seq_along(shortlist)], shortlist)
SL_LAB <- setNames(sapply(shortlist, function(v) VARIANTS[[v]]$label), shortlist)
# a size-matched random control for purity: items in random units, 20 draws
set.seed(1)
purity_random <- sapply(names(datasets), function(k) { d <- datasets[[k]]; mean(replicate(20, purity(sample(REF^2, nrow(d$X), TRUE), d$cls))) })
```

`r length(VARIANTS) - 2` distances and ingredients beyond the two controls were tested in `som_next_steps.Rmd`, on
`r length(datasets)` data sets, four map sizes and `r p$seeds` seeds, every
map judged with the Euclidean distance, with DTW and with its own. This page
keeps the `r p$n_short` that rank best on a composite of the measures that
judge a map's use (ARI, NMI, TE, KL, trustworthiness and continuity on each
data set, and the detection AUROC on the partial discharges), adds the best
variant on ARI of every data set, keeps the two controls, and shows them with
all the quality measures of the project, at `r REF` &times; `r REF` units
unless said otherwise.

<div class="box">

**The shortlist.** The rank is the mean of four blocks: the mean rank over `r length(RANK_ON)` measures on each of the
`r length(datasets)` data sets (`r paste(SHORT, collapse = ", ")`), and the rank on the detection test (1 = best of 17).
ARI is the mean over the data sets, AUROC the mean over the seeds.

```{r shortlist-table, echo = FALSE}
st <- rank_tab[rank_tab$variant %in% shortlist, ]
st$why <- sapply(st$variant, function(v) VARIANTS[[v]]$what)
st$star <- sapply(st$variant, function(v) { w <- names(ari_best)[ari_best == v]; if (length(w)) paste(SHORT[w], collapse = ", ") else "" })
blk <- round(st[, c(names(datasets), "detection")], 1); names(blk) <- c(SHORT, "Detection")
knitr::kable(data.frame(Variant = st$label, Rank = sprintf("%.1f", st$mean_rank), blk,
                        `ARI` = sprintf("%.2f", st$ari), `AUROC` = sprintf("%.2f", st$auroc), `Best ARI on` = st$star, check.names = FALSE),
             row.names = FALSE, align = "lrrrrrrrl")
```

What each one changes:

```{r shortlist-what, echo = FALSE, results = "asis"}
cat(sprintf("* **%s.** %s", st$label, st$why), sep = "\n")
```

Left out, with their mean rank: `r paste(sprintf("%s (%.1f)", rank_tab$label[!rank_tab$variant %in% shortlist], rank_tab$mean_rank[!rank_tab$variant %in% shortlist]), collapse = ", ")`.

</div>

# 1. Every measure, side by side

Pick a measure. Each row is a variant of the shortlist, each panel a data set;
the dot is the mean over `r p$seeds` seeds and the bar the range over the
seeds, so a difference inside the bar is seed-to-seed noise. The dotted line
marks the online SOM, the control.

```{r forest, fig.height = 5}
forest <- function() {
  keys <- names(datasets); nk <- length(keys); ypos <- rev(seq_along(shortlist))
  build <- function(me) lapply(keys, function(k) {
    x <- sapply(shortlist, function(v) { r <- val(k, v, REF, me); c(mean(r, na.rm = TRUE), min(r, na.rm = TRUE), max(r, na.rm = TRUE)) })
    list(m = I(x[1, ]), lo = I(x[2, ]), hi = I(x[3, ]), ref = mean_of(k, "euclid", REF, me),
         text = I(sprintf("<b>%s</b>, %s<br>%s: %s (seeds %s to %s)", SL_LAB, SHORT[[k]], MEAS[[me]]$label,
                          formatC(x[1, ], digits = 3, format = "fg", flag = "#"), formatC(x[2, ], digits = 3, format = "fg", flag = "#"), formatC(x[3, ], digits = 3, format = "fg", flag = "#"))))
  })
  B <- lapply(names(MEAS), build); names(B) <- names(MEAS)
  pl <- plot_ly(height = 420)
  lay <- list(margin = list(t = 50, b = 40, l = 10, r = 10), showlegend = FALSE, shapes = list())
  for (i in seq_len(nk)) {
    b <- B[[1]][[i]]; xa <- paste0("x", i); ya <- paste0("y", i)
    pl <- add_trace(pl, x = b$m, y = ypos, type = "scatter", mode = "markers", xaxis = xa, yaxis = ya, text = b$text, hovertemplate = "%{text}<extra></extra>",
                    error_x = list(type = "data", symmetric = FALSE, array = b$hi - b$m, arrayminus = b$m - b$lo, color = AXIS, thickness = 1.5, width = 0),
                    marker = list(color = unname(SL_COL), size = 11, line = list(color = "#ffffff", width = 1.5)))
    dom <- c((i - 1) / nk + 0.03, i / nk - 0.02)
    lay[[paste0("xaxis", if (i > 1) i else "")]] <- ax(SHORT[[keys[i]]], domain = dom, anchor = ya, zeroline = FALSE)
    lay[[paste0("yaxis", if (i > 1) i else "")]] <- ax(NULL, anchor = xa, tickvals = ypos, ticktext = if (i == 1) unname(SL_LAB) else rep("", length(ypos)),
                                                     range = c(0.4, length(ypos) + 0.6), showgrid = TRUE, ticksuffix = "  ")
    lay$shapes[[i]] <- list(type = "line", xref = xa, yref = ya, x0 = b$ref, x1 = b$ref, y0 = 0.4, y1 = length(ypos) + 0.6, line = list(color = INK2, width = 1.2, dash = "dot"))
  }
  buttons <- lapply(names(MEAS), function(me) {
    xs <- list(); ex <- list(); ts <- list(); shapes <- lay$shapes
    for (i in seq_len(nk)) { b <- B[[me]][[i]]; xs[[i]] <- b$m; ts[[i]] <- b$text
      ex[[i]] <- list(type = "data", symmetric = FALSE, array = b$hi - b$m, arrayminus = b$m - b$lo, color = AXIS, thickness = 1.5, width = 0)
      shapes[[i]]$x0 <- b$ref; shapes[[i]]$x1 <- b$ref }
    args2 <- list(shapes = shapes)
    for (i in seq_len(nk)) args2[[paste0("xaxis", if (i > 1) i else "", ".autorange")]] <- TRUE
    list(method = "update", label = paste0(MEAS[[me]]$label, if (MEAS[[me]]$higher) " \u00b7 higher is better" else " \u00b7 lower is better"),
         args = list(list(x = xs, text = ts, error_x = ex), args2))
  })
  lay$updatemenus <- list(menu(buttons, type = "dropdown", x = 0, y = 1.04))
  slim(do.call(chart, c(list(pl), lay)))
}
forest()
```

<p class="note">Purity rises with the number of units whatever the map learned: random units would score
`r paste(sprintf("%s %.2f", SHORT, purity_random), collapse = ", ")` on these sets at this size. EV, TE and KL
are judged within one size only; QE is left out because each map fits best in its own distance, and KL is
judged with DTW for every map (the common yardstick on series).</p>

# 2. The profile of each variant

Every measure scaled from the worst of the seventeen variants (0) to the best
(1) on the chosen data set: a line along the top is good everywhere, a dip is
the measure the variant pays with.

```{r profile, fig.height = 5.2}
profile <- function() {
  keys <- names(datasets); mids <- names(MEAS); nv <- length(shortlist); ncol <- 4; nrow <- ceiling(nv / ncol)
  scaled <- lapply(keys, function(k) sapply(mids, function(me) {
    vals <- sapply(vids, mean_of, ds = k, size = REF, me = me); if (!MEAS[[me]]$higher) vals <- -vals
    (vals - min(vals, na.rm = TRUE)) / (max(vals, na.rm = TRUE) - min(vals, na.rm = TRUE)) }))
  names(scaled) <- keys
  panels <- lapply(seq_along(shortlist), function(i) {
    v <- shortlist[i]; pl <- plot_ly()
    for (k in keys) {
      for (o in setdiff(vids, v)) pl <- add_trace(pl, x = mids, y = scaled[[k]][o, ], type = "scatter", mode = "lines", line = list(color = GRID, width = 1),
                                                 hoverinfo = "skip", showlegend = FALSE, visible = k == keys[1])
      pl <- add_trace(pl, x = mids, y = scaled[[k]][v, ], type = "scatter", mode = "lines+markers", line = list(color = SL_COL[[v]], width = 2.5),
                      marker = list(color = SL_COL[[v]], size = 7, line = list(color = "#ffffff", width = 1)), showlegend = FALSE, visible = k == keys[1],
                      text = sprintf("<b>%s</b>, %s<br>%s<br>%.2f of the way from the worst to the best of the 17", SL_LAB[[v]], SHORT[[k]], sapply(MEAS, `[[`, "label"), scaled[[k]][v, ]),
                      hovertemplate = "%{text}<extra></extra>")
    }
    pl
  })
  fig <- subplot(panels, nrows = nrow, shareX = TRUE, shareY = TRUE, margin = c(0.015, 0.015, 0.06, 0.06))
  n_per_panel <- length(vids) * length(keys)
  buttons <- lapply(keys, function(k) list(method = "restyle", label = SHORT[[k]], args = list(list(visible = rep(rep(keys == k, each = length(vids)), nv)))))
  ann <- lapply(seq_along(shortlist), function(i) {
    r <- (i - 1) %/% ncol; cidx <- (i - 1) %% ncol
    list(text = SL_LAB[[shortlist[i]]], xref = "paper", yref = "paper", x = (cidx + 0.5) / ncol, y = 1 - r / nrow + 0.005, xanchor = "center", yanchor = "bottom",
         showarrow = FALSE, font = list(size = 12, color = INK)) })
  lay <- list(updatemenus = list(menu(buttons, x = 0, y = 1.1)), annotations = ann, margin = list(t = 60, b = 50, l = 40, r = 10), showlegend = FALSE)
  for (i in seq_len(nv)) { lay[[paste0("xaxis", if (i > 1) i else "")]] <- ax(NULL, showgrid = FALSE, tickfont = list(size = 10, color = INK2), tickangle = -60)
                           lay[[paste0("yaxis", if (i > 1) i else "")]] <- ax(NULL, range = c(-0.08, 1.08), tickvals = c(0, 0.5, 1), tickfont = list(size = 10, color = INK2)) }
  slim(do.call(chart, c(list(fig), lay)))
}
profile()
```

<p class="note">0 is the worst and 1 the best of the seventeen variants on that measure and data set; the grey lines are the other sixteen.</p>

# 3. The maps

Seed 1, `r REF` &times; `r REF`. Three views of the same map: the **classes**
(each cell tinted with the class of most of the items it wins, white when it
wins none), the **U-matrix** (each cell shaded by the mean distance, in the
map's own distance, between its prototype and its neighbours' prototypes:
dark ridges are the walls between clusters), and the **hits** (how many items
each cell wins). Hover a cell for its numbers.

## Maps {.tabset .tabset-pills}

```{r maps, results = "asis"}
seq_ramp <- function(x, lo = "#cde2fb", hi = "#0d366b") { x <- (x - min(x)) / max(1e-12, diff(range(x))); grDevices::colorRampPalette(c(lo, hi))(101)[1 + round(100 * x)] }
map_views <- function(k) {
  d <- datasets[[k]]; col <- class_colours(d$cls); grid <- make_grid(REF, REF)
  W <- 640; H <- 560; pitch <- min((W - 20) / (REF + 1.1), (H - 50) / (REF * 0.87 + 0.8))   # pixels per lattice unit
  pl <- plot_ly(width = W, height = H); vis <- c()
  cp <- cell_paths(grid)
  for (v in shortlist) {
    f <- fits[[paste(k, v, REF, sep = "|")]]; if (is.null(f)) next
    spec <- VARIANTS[[v]]
    dV <- cross_dist(f$M, f$M, d, judge_args(d, "own", spec)); diag(dV) <- 0
    umat <- sapply(seq_len(grid$K), function(u) { nb <- which(grid$links[u, ] == 1); mean(dV[u, nb]) })
    counts <- table(factor(f$bmu, levels = seq_len(grid$K)), d$cls); hits <- rowSums(counts)
    main <- apply(counts, 1, function(r) if (sum(r)) colnames(counts)[which.max(r)] else "none")
    hover <- sprintf("unit %d: %d items, mostly %s<br>U-matrix %.3f<br>%s", seq_len(grid$K), hits, main, umat,
                     apply(counts, 1, function(r) paste(sprintf("%s %d", colnames(counts), r), collapse = ", ")))
    fills <- list(classes = cell_fill(f$bmu, d$cls, grid$K, col), umatrix = seq_ramp(umat), hits = seq_ramp(hits, "#fcfcfb", "#184f95"))
    for (view in names(fills)) {
      pl <- add_trace(pl, x = grid$pos[, 1], y = grid$pos[, 2], type = "scatter", mode = "markers",
                      marker = list(symbol = "hexagon", size = 0.72 * pitch, color = fills[[view]], line = list(color = "#c3c2b7", width = 0.6)),
                      text = hover, hoverinfo = "text", name = paste(v, view), visible = v == shortlist[1] && view == "classes")
      vis <- c(vis, paste(v, view))
    }
  }
  pl <- add_trace(pl, x = round(cp$x, 3), y = round(cp$y, 3), type = "scatter", mode = "lines", line = list(color = "#c3c2b7", width = 0.6), hoverinfo = "skip", name = "outline")
  vis <- c(vis, "outline")
  variants_here <- unique(sub(" (classes|umatrix|hits)$", "", vis[vis != "outline"]))
  # two menus would fight over 'visible', so one dropdown carries variant x view
  VIEWS <- c(classes = "classes", umatrix = "U-matrix", hits = "hits")
  buttons <- list()
  for (v in variants_here) for (view in names(VIEWS))
    buttons[[length(buttons) + 1]] <- list(method = "restyle", label = paste0(SL_LAB[[v]], ": ", VIEWS[[view]]),
                                           args = list(list(visible = vis == paste(v, view) | vis == "outline")))
  chart(pl, xaxis = list(visible = FALSE, range = c(0.2, REF + 1.3), fixedrange = TRUE), yaxis = list(visible = FALSE, range = c(0.2, REF * 0.87 + 1), fixedrange = TRUE, scaleanchor = "x"),
        showlegend = FALSE, updatemenus = list(menu(buttons, type = "dropdown", x = 0, y = 1.02)), margin = list(t = 40, b = 10, l = 10, r = 10)) |> slim()
}
for (k in names(datasets)) { cat("\n\n### ", SHORT[[k]], "\n\n", sep = ""); print(tagList(map_views(k))); cat("\n\n") }
```

<p class="note">Class colours: `r paste(sprintf("<span style='color:%s'>&#9632;</span> %s", unname(class_colours(datasets[[1]]$cls)), levels(datasets[[1]]$cls)), collapse = "  ")` for CBF,
and in the same order for the other sets. The U-matrix and the hits use one blue ramp, light to dark.</p>

# 4. Against the control, seed by seed

On how many of the `r p$seeds` seeds each variant beat the online SOM with
the same start and order of items, minus the seeds it lost, for every
measure. Blue: better on more seeds than worse; red: the other way.

```{r wins, fig.height = 4.2}
wins_hm <- function() {
  keys <- names(datasets); vs <- setdiff(shortlist, "euclid")
  Z <- lapply(keys, function(k) t(sapply(vs, function(v) sapply(names(MEAS), function(me) {
    a <- val(k, v, REF, me); b <- val(k, "euclid", REF, me); n <- min(length(a), length(b)); d <- if (MEAS[[me]]$higher) a[1:n] - b[1:n] else b[1:n] - a[1:n]
    sum(d > 0, na.rm = TRUE) - sum(d < 0, na.rm = TRUE) }))))
  H <- lapply(keys, function(k) t(sapply(vs, function(v) sapply(names(MEAS), function(me) {
    a <- val(k, v, REF, me); b <- val(k, "euclid", REF, me)
    sprintf("<b>%s</b>, %s<br>%s: %.3g against %.3g for the online SOM", SL_LAB[[v]], SHORT[[k]], MEAS[[me]]$label, mean(a, na.rm = TRUE), mean(b, na.rm = TRUE)) }))))
  pl <- plot_ly(x = sapply(MEAS, `[[`, "id"), y = unname(SL_LAB[vs]), z = Z[[1]], text = H[[1]], type = "heatmap", hoverinfo = "text",
                colorscale = list(c(0, "#e34948"), c(0.25, "#f3aeac"), c(0.5, "#f0efec"), c(0.75, "#9ec5f4"), c(1, "#2a78d6")), zmin = -p$seeds, zmax = p$seeds, xgap = 3, ygap = 3,
                colorbar = list(title = list(text = "seeds better minus worse", font = list(size = 12, color = INK2)), tickvals = c(-p$seeds, 0, p$seeds), thickness = 12, outlinewidth = 0))
  ann <- lapply(Z, function(z) unlist(lapply(seq_len(nrow(z)), function(i) lapply(seq_len(ncol(z)), function(j)
    list(x = names(MEAS)[j], y = SL_LAB[[vs[i]]], text = sprintf("%+d", z[i, j]), showarrow = FALSE,
         font = list(size = 11, color = if (abs(z[i, j]) >= 4) "#ffffff" else INK)))), recursive = FALSE))
  buttons <- lapply(seq_along(keys), function(i) list(method = "update", label = SHORT[[keys[i]]],
                                                      args = list(list(z = list(Z[[i]]), text = list(H[[i]])), list(annotations = ann[[i]]))))
  chart(pl, xaxis = ax(NULL, showgrid = FALSE), yaxis = ax(NULL, autorange = "reversed", showgrid = FALSE, ticksuffix = "  "),
        annotations = ann[[1]], updatemenus = list(menu(buttons, x = 0, y = 1.04)), margin = list(t = 50, l = 200)) |> slim()
}
wins_hm()
```

# 5. Detection and cost

```{r detect, fig.height = 3.8}
BASE_LAB <- c(base_peak = "Control: loudness threshold", base_1nn_euclid = "Control: nearest noise item, Euclidean",
              base_1nn_cosine = "Control: nearest noise item, cosine", base_1nn_dtw = "Control: nearest noise item, DTW")
dk <- det[det$variant %in% c(shortlist, names(BASE_LAB)), ]
lab <- function(v) if (v %in% names(SL_LAB)) SL_LAB[[v]] else BASE_LAB[[v]]
da <- aggregate(AUROC ~ variant, dk, mean); da <- da[order(-da$AUROC), ]
pl <- plot_ly(height = 340)
for (v in da$variant) { d <- dk[dk$variant == v, ]; base <- v %in% names(BASE_LAB)
  pl <- add_trace(pl, x = d$AUROC, y = rep(lab(v), nrow(d)), type = "scatter", mode = "markers", name = lab(v), showlegend = FALSE,
                  marker = list(color = if (base) MUTED else SL_COL[[v]], symbol = if (base) "diamond" else "circle", size = if (base) 12 else 10, line = list(color = "#ffffff", width = 1.5)),
                  text = sprintf("<b>%s</b>%s<br>AUROC %.3f", lab(v), if (base) "" else sprintf(", seed %d", d$seed), d$AUROC), hovertemplate = "%{text}<extra></extra>") }
chart(pl, xaxis = ax("AUROC of the alarm (maps trained on noise only)", range = c(0.4, 1.02)),
      yaxis = ax(NULL, categoryorder = "array", categoryarray = rev(sapply(da$variant, lab)), ticksuffix = "  "),
      shapes = list(list(type = "line", x0 = 0.5, x1 = 0.5, yref = "paper", y0 = 0, y1 = 1, line = list(color = AXIS, dash = "dot"))), margin = list(t = 10, l = 240)) |> slim()
```

```{r cost}
secs <- sapply(shortlist, function(v) sapply(names(datasets), function(k) mean(bench$seconds[bench$dataset == k & bench$variant == v & bench$size == REF & bench$judge == "own"])))
cost <- data.frame(Variant = unname(SL_LAB), t(round(secs, 2)), check.names = FALSE)
names(cost)[-1] <- paste("Seconds per map,", SHORT)
cost$`Relative to the online SOM` <- sprintf("%.0f x", colMeans(secs) / colMeans(secs)[["euclid"]])
tab(cost, search = FALSE)
```

# 6. All numbers

Means and seed-to-seed standard deviations of every measure for the shortlist,
at every size; filter with the menus.

```{r numbers}
rows <- do.call(rbind, lapply(names(datasets), function(k) do.call(rbind, lapply(SIZES, function(s) do.call(rbind, lapply(shortlist, function(v) {
  cells <- sapply(names(MEAS), function(me) { x <- val(k, v, s, me); if (!length(x) || all(is.na(x))) "" else sprintf("%.3f &plusmn; %.3f", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE)) })
  data.frame(`Data set` = SHORT[[k]], Map = sprintf("%d x %d", s, s), Variant = SL_LAB[[v]], t(cells), check.names = FALSE, stringsAsFactors = FALSE)
}))))))
names(rows)[-(1:3)] <- sapply(MEAS, `[[`, "id")
filtered_table(rows, c("Data set", "Map", "Variant"), id = "shortlist-numbers", escape = FALSE, page_length = length(shortlist) * 2)
```

<div class="box">

* **The shortlist is a ranking, not a verdict.** With `r p$seeds` seeds, only
  `r p$seeds` of `r p$seeds` is a statement (two-sided sign test p = `r signif(2 / 2^p$seeds, 2)`);
  the heatmap of Section 4 shows how few cells reach it.
* **Each measure answers one question.** EV says how close the prototypes are,
  TE and KL whether the map is in order, trustworthiness and continuity whether
  the neighbourhoods are the data's, purity, ARI and NMI whether the classes
  came out, the AUROC whether a discharge trips the alarm. The profile of
  Section 2 shows what each variant pays for what it gains.
* **These are synthetic stand-ins.** Put `som_bench/data/datasets.rds` next to
  the code and both reports use the real CBF, partial discharges and GunPoint.

</div>
