---
title: "Where to go next with the SOM"
subtitle: "A brainstorm for time series, partial discharges and movement: what the benchmarks taught, the ideas worth a seed-by-seed test, and a demonstrator of nine distances and six training ingredients on four map sizes, down to the last bit of precision"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: hide
    mathjax: null
params:
  seeds:
    label: "Seeds per variant, data set and map size"
    value: 5
    input: slider
    min: 2
    max: 30
    step: 1
  rlen:
    label: "Passes over the data"
    value: 10
    input: slider
    min: 2
    max: 50
    step: 1
  cores:
    label: "CPU cores to use (0 = all but two)"
    value: 0
    input: numeric
  mpfr_bits:
    label: "Bits of the MPFR reference"
    value: 256
    input: numeric
  quick:
    label: "Quick mode: 5 x 5 and 7 x 7 only, a small ladder, a separate cache (for development)"
    value: no
    input: checkbox
---

```{css, echo = FALSE}
div.main-container { max-width: 1380px; }
body { font-size: 16px; line-height: 1.6; color: #0b0b0b; }
.box { border-left: 4px solid #2a78d6; background: #f3f7fd; padding: 10px 18px; margin: 16px 0 22px; }
.box p:last-child, .box ul:last-child { margin-bottom: 0; }
.warn { border-left: 4px solid #eb6834; background: #fdf4ee; padding: 10px 18px; margin: 16px 0 22px; }
.note { color: #52514e; font-size: 14.5px; }
.chip { display: inline-block; padding: 1px 9px; border-radius: 10px; font-size: 12.5px; font-weight: 600; white-space: nowrap; }
.chip.A { background: #dcefdc; color: #0b5d0b; } .chip.B { background: #e3eefc; color: #184f95; } .chip.C { background: #fbeccb; color: #7a4f00; }
.chip.exact { background: #dcefdc; color: #0b5d0b; } .chip.metaphor { background: #efeeea; color: #52514e; }
table.dataTable { font-size: 13.5px; } table.dataTable td { vertical-align: top; }
details { margin: 6px 0 10px; } details > summary { cursor: pointer; color: #184f95; }
td details > summary { font-size: 13px; }
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center", out.width = "100%", comment = "#>")
options(knitr.kable.NA = "")
library(plotly); library(DT); library(htmltools)
```

```{r code, results = "hide"}
# the compiled core and the R files of next_steps/ (every file is shown in full in Section 4)
if (!exists("params")) {                                   # running the chunks one by one
  header <- rmarkdown::yaml_front_matter("som_next_steps.Rmd")$params
  params <- lapply(header, function(x) if (is.list(x)) x$value else x)
}
p <- params
cores <- if (p$cores > 0) p$cores else max(1, parallel::detectCores() - 2)
needed <- c("Rcpp", "digest", "jsonlite", "htmltools", "xfun", "plotly", "DT", "Rmpfr")
missing <- needed[!vapply(needed, requireNamespace, TRUE, quietly = TRUE)]
if (length(missing)) stop("please install: ", paste(missing, collapse = ", "))
Sys.setenv(PKG_LIBS = "-lquadmath -lmpfr -lgmp")           # the 128-bit and the MPFR arithmetic of the precision ladder
Rcpp::sourceCpp("next_steps/ns_core.cpp", cacheDir = "next_steps/.cpp_cache")
Rcpp::sourceCpp("next_steps/ns_mpfr.cpp", cacheDir = "next_steps/.cpp_cache")
for (f in c("ns_data.R", "ns_maps.R", "ns_quality.R", "ns_bench.R", "ns_precision.R", "ns_report.R")) source(file.path("next_steps/R", f))
```

```{r run-all, include = FALSE}
# Everything is computed here, before the first line of text, and cached in next_steps/cache under a key
# made from the code and the settings. The first knit takes about two hours of CPU time (half of it
# soft-DTW); next_steps/run_benchmark.R computes the same cache outside the knit, on all cores.
QUICK <- isTRUE(p$quick)
SIZES <- if (QUICK) c(5, 7) else c(5, 7, 10, 14)
REF <- if (10 %in% SIZES) 10 else max(SIZES)                   # the size the text quotes
CACHE_DIR <- if (QUICK) "next_steps/cache_quick" else "next_steps/cache"
seeds <- seq_len(p$seeds)
datasets <- load_datasets()
time_all <- system.time(res <- compute_all(datasets, VARIANTS, SIZES, seeds, p$rlen, cores, dir = CACHE_DIR))[["elapsed"]]
bench <- res$bench; fits <- res$fits; det <- res$detection
n_maps <- nrow(unique(bench[, c("dataset", "variant", "size", "seed")]))
cpu_min <- sum(unique(bench[, c("dataset", "variant", "size", "seed", "seconds")])$seconds) / 60

# the precision ladder: the tutorial's small CBF (with the Rmpfr check of the MPFR core) and the PD stand-in
LADDER <- if (QUICK) list(cbf = list(size = 5, rlen = 4), pd = list(size = 5, rlen = 1), pd_raw = list(size = 5, rlen = 1)) else list(cbf = list(size = 7, rlen = 20), pd = list(size = 7, rlen = 4), pd_raw = list(size = 7, rlen = 4))
ladder <- compute_ladder(datasets, p$mpfr_bits, LADDER, dir = CACHE_DIR)
small_cbf <- ladder$small_cbf

# the verified idea catalogue of the brainstorm (next_steps/ideas/ideas.json), if it is there
ideas <- if (file.exists("next_steps/ideas/ideas.json")) jsonlite::fromJSON("next_steps/ideas/ideas.json", simplifyVector = FALSE) else NULL

# ------------------------------------------------------------------------------- helpers for the text
JUDGE_LAB <- c(euclid = "Euclidean", dtw = "DTW", own = "own distance")
VIEWS <- list(                                             # measure, judge, how the advantage over the control is taken
  list(id = "ari_own",  m = "ARI",    j = "own",    higher = TRUE,  kind = "diff", label = "ARI, each map with its own distance"),
  list(id = "ari_dtw",  m = "ARI",    j = "dtw",    higher = TRUE,  kind = "diff", label = "ARI, all read with DTW"),
  list(id = "ari_eu",   m = "ARI",    j = "euclid", higher = TRUE,  kind = "diff", label = "ARI, all read with the Euclidean distance"),
  list(id = "nmi_own",  m = "NMI",    j = "own",    higher = TRUE,  kind = "diff", label = "NMI, own distance"),
  list(id = "pur_own",  m = "purity", j = "own",    higher = TRUE,  kind = "diff", label = "Purity, own distance"),
  list(id = "qe_eu",    m = "QE",     j = "euclid", higher = FALSE, kind = "rel",  label = "QE, all judged with the Euclidean distance"),
  list(id = "qe_dtw",   m = "QE",     j = "dtw",    higher = FALSE, kind = "rel",  label = "QE, all judged with DTW"),
  list(id = "ev_eu",    m = "EV",     j = "euclid", higher = TRUE,  kind = "diff", label = "EV (explained variance), Euclidean"),
  list(id = "te_own",   m = "TE",     j = "own",    higher = FALSE, kind = "pp",   label = "TE, own distance"),
  list(id = "te_dtw",   m = "TE",     j = "dtw",    higher = FALSE, kind = "pp",   label = "TE, all read with DTW"),
  list(id = "kl_eu",    m = "KL",     j = "euclid", higher = FALSE, kind = "rel",  label = "KL (Kaski-Lagus), Euclidean"),
  list(id = "kl_dtw",   m = "KL",     j = "dtw",    higher = FALSE, kind = "rel",  label = "KL (Kaski-Lagus), DTW"),
  list(id = "tp_own",   m = "TP",     j = "own",    higher = FALSE, kind = "diff", label = "|TP| (topographic product), own distance"),
  list(id = "used_own", m = "used",   j = "own",    higher = TRUE,  kind = "pp",   label = "Share of units used, own distance"),
  list(id = "trust",    m = "trust",  j = "euclid", higher = TRUE,  kind = "diff", label = "Trustworthiness (Euclidean neighbourhoods)"),
  list(id = "cont",     m = "cont",   j = "euclid", higher = TRUE,  kind = "diff", label = "Continuity (Euclidean neighbourhoods)"),
  list(id = "secs",     m = "seconds", j = "own",   higher = FALSE, kind = "rel",  label = "Seconds to train"))
names(VIEWS) <- sapply(VIEWS, `[[`, "id")
val <- function(ds, v, size, m, j) {                       # one measure of one map, every seed in order
  b <- bench[bench$dataset == ds & bench$variant == v & bench$size == size & bench$judge == j, ]
  b <- b[order(b$seed), ]; x <- b[[m]]; if (m == "TP") abs(x) else x
}
mean_of <- function(ds, v, size, view) mean(val(ds, v, size, VIEWS[[view]]$m, VIEWS[[view]]$j), na.rm = TRUE)
adv_of <- function(ds, v, size, view) {                    # the variant against its family's control, seed by seed
  vw <- VIEWS[[view]]; ctrl <- VARIANTS[[v]]$family
  a <- val(ds, v, size, vw$m, vw$j); b <- val(ds, ctrl, size, vw$m, vw$j)
  n <- min(length(a), length(b)); a <- a[seq_len(n)]; b <- b[seq_len(n)]
  adv <- switch(vw$kind, diff = if (vw$higher) a - b else b - a, pp = 100 * (if (vw$higher) a - b else b - a),
                rel = 100 * (if (vw$higher) (a - b) / b else (b - a) / b))
  data.frame(seed = seq_len(n), v = a, c = b, adv = adv)
}
wins_of <- function(ds, v, size, view) { a <- adv_of(ds, v, size, view)$adv; c(better = sum(a > 0, na.rm = TRUE), worse = sum(a < 0, na.rm = TRUE), n = sum(!is.na(a))) }
# the same against any other map (for the DTW-SOM against the online SOM, say)
wins_vs <- function(ds, v, ctrl, size, view) {
  vw <- VIEWS[[view]]; a <- val(ds, v, size, vw$m, vw$j); b <- val(ds, ctrl, size, vw$m, vw$j)
  n <- min(length(a), length(b)); a <- a[seq_len(n)]; b <- b[seq_len(n)]
  d <- if (vw$higher) a - b else b - a
  c(better = sum(d > 0, na.rm = TRUE), worse = sum(d < 0, na.rm = TRUE), n = sum(!is.na(d)))
}
# words for a count of seeds: 'better on 4 of 5 seeds, worse on 1'
tally <- function(w) if (w[["n"]] == 0) "not run" else sprintf("better on %d of %d seeds, worse on %d", w[["better"]], w[["n"]], w[["worse"]])
tally_short <- function(w) if (w[["n"]] == 0) "not run" else sprintf("%d of %d", w[["better"]], w[["n"]])
judge_word <- function(w) { n <- w[["n"]]; if (n == 0) "not run" else if (w[["better"]] >= 0.8 * n && w[["better"]] > w[["worse"]]) "clearly better" else if (w[["worse"]] >= 0.8 * n && w[["worse"]] > w[["better"]]) "clearly worse" else "level" }
# one phrase per data set: 'CBF clearly worse (0 of 5), partial discharges level (3 of 5), gestures ...'
by_ds <- function(v, view, size = REF, ctrl = NULL) paste(sapply(names(datasets), function(k) {
  w <- if (is.null(ctrl)) wins_of(k, v, size, view) else wins_vs(k, v, ctrl, size, view)
  sprintf("%s %s (%s)", SHORT[[k]], judge_word(w), tally_short(w)) }), collapse = ", ")
# on how many data sets a variant was clearly better / clearly worse than its control on a view
count_ds <- function(v, view, size = REF) { j <- sapply(names(datasets), function(k) judge_word(wins_of(k, v, size, view))); c(better = sum(j == "clearly better"), worse = sum(j == "clearly worse"), n = length(j)) }
best_variant <- function(ds, size, view, among = names(VARIANTS)) {
  m <- sapply(among, mean_of, ds = ds, size = size, view = view)
  m <- m[is.finite(m)]
  if (!VIEWS[[view]]$higher) m <- -m
  list(id = names(m)[which.max(m)], label = VARIANTS[[names(m)[which.max(m)]]]$label, value = abs(m[which.max(m)]))
}
SHORT <- c(cbf = "CBF", pd = "Partial discharges", gestures = "Gestures", gunpoint = "GunPoint")
SHORT <- SHORT[names(datasets)]
det_mean <- function(v) mean(det$AUROC[det$variant == v])
v_lab <- function(v) VARIANTS[[v]]$label
nn_floor <- sapply(datasets, function(d) sapply(c("euclid", "dtw"), function(judge) {
  Dm <- cross_dist(d$X, d$X, d, judge_args(d, judge)); diag(Dm) <- Inf
  mean(d$cls[max.col(-Dm, ties.method = "first")] == d$cls) }))
```

```{r headline-numbers, include = FALSE}
# the numbers of the box below, all from the results of this knit (REF x REF maps unless said otherwise)
pd_best_ari <- best_variant("pd", REF, "ari_own")
pd_eu_ari   <- mean_of("pd", "euclid", REF, "ari_own"); pd_dtw_ari <- mean_of("pd", "dtw", REF, "ari_own")
ge_best_ari <- best_variant("gestures", REF, "ari_own"); ge_eu_ari <- mean_of("gestures", "euclid", REF, "ari_own")
det_best <- names(which.max(sapply(names(VARIANTS), det_mean)))
det_corona_eu <- mean(det$AUROC_corona[det$variant == "euclid"])
# the two temperatures against their controls, judged with DTW for soft-DTW (the control's own distance)
soft_fit <- by_ds("softdtw", "qe_dtw"); soft_cls <- by_ds("softdtw", "ari_dtw"); soft_te <- by_ds("softdtw", "te_dtw")
ann_te <- by_ds("anneal", "te_own"); ann_qe <- by_ds("anneal", "qe_eu"); ann_cls <- by_ds("anneal", "ari_own")
temps_won <- sum(sapply(c("softdtw", "anneal", "dtw_anneal"), function(v) sum(sapply(c("qe_dtw", "qe_eu", "te_own", "ari_own", "kl_dtw", "kl_eu"), function(vw) count_ds(v, vw)["better"]))))
# the distances against the ONLINE SOM on the classes (the DTW-SOM is its own family's control)
THR <- ceiling(0.8 * p$seeds)                              # 'clearly': at least this many seeds, and more better than worse
dist_vs_eu <- function(k) { v <- setdiff(names(VARIANTS)[sapply(VARIANTS, `[[`, "group") == "distance"], "euclid")
  w <- sapply(v, function(x) wins_vs(k, x, "euclid", REF, "ari_own")); v[w["better", ] >= THR & w["better", ] > w["worse", ]] }
temps_n <- 3 * 6 * length(datasets)
clear_pd <- dist_vs_eu("pd"); clear_ge <- dist_vs_eu("gestures"); clear_cbf <- dist_vs_eu("cbf")
size_flip  <- sapply(names(datasets), function(k) sapply(SIZES, function(s) best_variant(k, s, "ari_own")$id))
LD <- c(cbf = "small CBF", pd = "PD stand-in (asinh)", pd_raw = "PD stand-in, raw pressure")
lad_tab <- do.call(rbind, lapply(names(LD), function(k) cbind(data = k, ladder[[k]]$table)))
fp32_first <- lad_tab[lad_tab$precision == "fp32" & lad_tab$mode == "euclid", c("data", "first_flip", "flips", "drift")]
fp64_flips <- lad_tab[lad_tab$precision == "fp64", c("data", "mode", "flips", "drift")]
gap_small  <- sapply(ladder[names(LD)], function(l) c(fp32 = mean(l$gap < 2^-23, na.rm = TRUE), fp64 = mean(l$gap < 2^-52, na.rm = TRUE), min = min(l$gap, na.rm = TRUE)))
raw32 <- lad_tab[lad_tab$data == "pd_raw" & lad_tab$mode == "euclid" & lad_tab$precision == "fp32", ]
raw64 <- lad_tab[lad_tab$data == "pd_raw" & lad_tab$mode == "euclid" & lad_tab$precision == "fp64", ]
raw32d <- lad_tab[lad_tab$data == "pd_raw" & lad_tab$mode == "dtw" & lad_tab$precision == "fp32", ]
flip_words <- function(flips, S) if (flips == 0) sprintf("the same winner at every one of the %s steps", format(S, big.mark = ",")) else sprintf("another winner at %d of the %s steps", flips, format(S, big.mark = ","))
first_words <- function(r, S) if (is.na(r$first_flip)) sprintf("never chose a different winner in %s steps either, and ended within %s of the reference", format(S, big.mark = ","), formatC(r$drift, digits = 1, format = "e")) else sprintf("first chose a different winner at step %d and then went its own way (prototypes %s apart at the end)", r$first_flip, formatC(r$drift, digits = 1, format = "e"))
e1 <- function(x) formatC(x, digits = 1, format = "e")
mp_name    <- sprintf("MPFR %d-bit", p$mpfr_bits)
mpfr_secs  <- lad_tab$seconds[lad_tab$precision == mp_name & lad_tab$data == "cbf" & lad_tab$mode == "euclid"]
fp64_secs  <- lad_tab$seconds[lad_tab$precision == "fp64" & lad_tab$data == "cbf" & lad_tab$mode == "euclid"]
rmpfr_row  <- lad_tab[grepl("Rmpfr", lad_tab$precision), ]
```

The project built a SOM from scratch, made it a DTW-SOM, and tested fourteen
variants on twelve data sets, plus chaos, rings and tearing maps. The growing
DTW-SOM was the best of its family and DTW the biggest single lever; chaos
and tearing lost to the plain controls that do the same job. This page asks
**what to build next** for two uses: **detecting and classifying partial
discharges** from their waveforms, and **classifying movements** from sensor
series. It has three parts:

1. **The ideas** (Sections 1 to 3): what the benchmarks already decided, a
   catalogue of improvements gathered from seven angles (distances, physics,
   the structure of the map, partial-discharge engineering, movement
   recognition, numerical precision, and evaluation), each with the plain
   control it must beat seed by seed, and a bridge from physics to the code.
2. **A demonstrator** (Sections 4 to 7): the online SOM rewritten once, with
   nine distances in the winner choice *and* in the update, six training
   ingredients that each test one idea, four map sizes, all the project's
   measures plus trustworthiness, continuity and detection AUROC, on
   `r length(datasets)` data sets (`r paste(sapply(datasets, "[[", "title"), collapse = "; ")`),
   `r p$seeds` seeds each, every variant judged against
   its family's control with the same start and order of items.
3. **The precision ladder** (Section 8): the same map in fp32, fp64, 80-bit,
   128-bit and `r p$mpfr_bits`-bit MPFR arithmetic, step by step.

<div class="box">
**The answer in short** (`r p$seeds` seeds, `r p$rlen` passes, `r REF` &times; `r REF` maps unless said otherwise; the synthetic stand-ins of Section 2)

* **The distance is the lever, and it must sit in the update.** On the
  partial-discharge stand-in the classes were found best by the
  `r pd_best_ari$label` (ARI `r f2(pd_best_ari$value)` against `r f2(pd_eu_ari)`
  for the online SOM and `r f2(pd_dtw_ari)` for the DTW-SOM); on the gestures by
  the `r ge_best_ari$label` (`r f2(ge_best_ari$value)` against `r f2(ge_eu_ari)`).
  Against the online SOM with the same seeds, the distances that found the
  classes better on at least `r THR` of `r p$seeds` seeds were
  `r if (length(clear_pd)) paste(sapply(clear_pd, v_lab), collapse = ", ") else "none"` on the
  partial discharges, `r if (length(clear_ge)) paste(sapply(clear_ge, v_lab), collapse = ", ") else "none"`
  on the gestures and `r if (length(clear_cbf)) paste(sapply(clear_cbf, v_lab), collapse = ", ") else "none"`
  on CBF. Manhattan, Chebyshev, cosine, correlation and the shape-based
  distance each change the winner for a reason, and the update stays a
  straight line towards the item; only DTW, soft-DTW and SBD change what the
  item looks like to the prototype.
* **Detection is a different question from classification.** With a map
  trained on noise alone and the distance to the winner as the alarm, the
  `r v_lab(det_best)` map told discharges from noise best (AUROC
  `r f2(det_mean(det_best))`, online SOM `r f2(det_mean("euclid"))`).
  `r if (det_corona_eu < 0.5) sprintf("On corona alone the Euclidean alarm scores %s: a quiet corona pulse is *closer* to the noise map than noise itself, because a Euclidean distance is mostly loudness.", f2(det_corona_eu)) else sprintf("On corona alone the Euclidean alarm scores %s.", f2(det_corona_eu))`
  An amplitude-blind distance asks about the shape instead, and that is what
  detection needs.
* **Physics gives two exact temperatures and several metaphors.** Soft-DTW is
  the free energy of all alignments at a temperature gamma, and its gradient
  is the expected alignment that the update uses; deterministic annealing
  puts a temperature on the winner. Both are principled where the chaotic
  clocks were not, and both were
  tested here against the plain map that does the same job. Soft-DTW against
  the DTW-SOM, both judged with DTW, on fit: `r soft_fit`; on the classes:
  `r soft_cls`. The annealed winner against the online SOM, on order (TE):
  `r ann_te`; on fit: `r ann_qe`; on the classes: `r ann_cls`.
  `r if (temps_won == 0) "Neither temperature was clearly better than its control on any data set and any measure at this size: a principled knob is still a knob, and the test that decides is the one inside the growing DTW-SOM on the real waveforms (Section 9)." else sprintf("The temperatures were clearly better than their controls in %d of %d data-set-and-measure comparisons at this size (three maps, six measures); the rest were level or worse. The test that decides is the one inside the growing DTW-SOM on the real waveforms (Section 9).", temps_won, temps_n)`
* **The map size changes who wins.** Read the charts of Section 6 at four
  sizes before choosing: the best distance for the classes on the
  partial-discharge stand-in was `r paste(unique(sapply(size_flip[, "pd"], v_lab)), collapse = " / ")`
  from 5 &times; 5 to 14 &times; 14, and every fit measure falls with the number of
  units whatever the map does (the project's trap, Section 4 of the metrics page).
* **Precision: fp64 is the right stop.** With the same seed, the fp64
  Euclidean map chose `r flip_words(fp64_flips$flips[fp64_flips$data == "cbf" & fp64_flips$mode == "euclid"], ladder$cbf$S)`
  on the small CBF as the `r p$mpfr_bits`-bit MPFR run, and its prototypes
  ended within `r e1(max(fp64_flips$drift[fp64_flips$data == "cbf"]))` of
  it; on the PD stand-in, `r flip_words(fp64_flips$flips[fp64_flips$data == "pd" & fp64_flips$mode == "euclid"], ladder$pd$S)`.
  fp32 `r first_words(fp32_first[fp32_first$data == "cbf", ], ladder$cbf$S)`.
  The closest contest of the PD run had a relative gap of
  `r e1(gap_small["min", "pd"])` between the first and the second distance,
  `r if (gap_small["fp32", "pd"] == 0) "never inside fp32's resolution" else sprintf("inside fp32's resolution in %s of the steps", pct(gap_small["fp32", "pd"]))`
  and `r if (gap_small["fp64", "pd"] == 0) "never inside fp64's" else sprintf("inside fp64's in %s", pct(gap_small["fp64", "pd"]))`.
  The one case built to fail is raw pressure without the asinh scale, four
  decades in one squared distance: there the fp32 Euclidean map
  `r if (is.na(raw32$first_flip)) sprintf("still chose the reference's winner at every step, within %s", e1(raw32$drift)) else sprintf("chose another winner at %d of %s steps, the first at step %d", raw32$flips, format(ladder$pd_raw$S, big.mark = ","), raw32$first_flip)`,
  the fp32 DTW map `r if (is.na(raw32d$first_flip)) "did the same" else sprintf("first chose another winner at step %d", raw32d$first_flip)`,
  and fp64 `r if (raw64$flips == 0) "agreed with the reference at every step" else sprintf("differed at %d steps", raw64$flips)`. MPFR is a reference oracle
  (`r sprintf("%.1f", mpfr_secs)` s in C for what fp64 does in
  `r sprintf("%.2f", fp64_secs)` s), not a training tool; the precision
  question that matters for a field detector is the other way round, fp32 or
  fixed point on a microcontroller, with fp64 as the oracle.
* **What to build first** (Section 9), in the order the brainstorm's
  skeptics and synthesizer settled on: read the cached maps on held-out
  items with per-class confusion, detection AUROC and a sign test over the
  seeds, before any new map; then the representation ladder for partial
  discharges on the real waveforms (pulse features, envelope and
  log-spectrum, loudness as a separate coordinate, a loudness-only control);
  then shift rather than warp (SBD, or a 2 to 3% band) against the 10%
  DTW-SOM; then Heskes' energy-consistent winner at a smaller start radius;
  and for movement, dependent against independent multichannel DTW under
  the 1-NN DTW ceiling. Tier B: soft-DTW's barycentre, LVQ fine-tuning,
  U*-matrix clusters, TWED and MSM. Not first: chaos, tearing of any kind,
  MPFR training, momentum, the torus, conscience.
</div>

# 1. What the benchmarks already decided

Seven reports tested every change one at a time, seed by seed, against the
plain map that does the same job. These are the findings that bind any new
idea, with the report they come from:

```{r lessons}
lessons <- data.frame(
  Lesson = c("DTW belongs in the update, not only in the winner choice",
             "DTW or not depends on the data more than any variant does",
             "The window is a property of the data",
             "Fit and order pull against each other; growing maps won on both",
             "Chaos never beat the plain rival that does the same job",
             "Tearing never beat a smaller start radius",
             "Part of DTW's gain is in reading the map, not in training it",
             "No single score decides; compare only same size, same data, same distance",
             "Prototypes lose their timing under DTW; births from the data give it back"),
  Evidence = c("DTW for the winner only gave the worst fit of all (QE higher than the DTW-SOM on all series sets).",
               "ARI better by more than 0.1 on CBF and Synthetic Control, worse by more than 0.1 on the solar I-V curves, where the position of a feature is the information.",
               "CBF: ARI 0.24 at a 10% window, 0.92 at 20%; solar I-V curves better at 5%.",
               "Batch: best QE, worse TE and KL; CHAOSOM: the opposite; growing SOM and growing DTW-SOM: lower QE, TE and KL on most sets.",
               "Restarts at even steps did what CHAOSOM's spikes did; the smooth long-run output did what the chaotic units did; chaotic units hurt everywhere (their memory, not the chaos).",
               "On 18 data sets and 3 sizes, tearing was never better than all three radius maps on QE and TE together; rejoining or mending did not rescue it.",
               "On the partial discharges the online SOM mapped and clustered with DTW came close to the DTW-SOM's ARI.",
               "QE, EV and KL fall with more units; TE is near 0 on tiny maps; purity rises with units; DTW is never above the Euclidean distance.",
               "Timing error 3.6 to 2.0 time points on CBF; the floating neighbourhood gives back the order the births cost."),
  Where = c("Tutorial 11a; benchmark", "Benchmark 10", "Benchmark 8", "Benchmark 7, 10", "Tutorial 13; chaotic units", "Tearing lab; tear and join",
            "Benchmark 10", "Metrics; benchmark 4", "Tutorial 12"),
  `What it means here` = c("Every new distance in Section 4 comes with its own update step.",
                           "The demonstrator runs every distance on three kinds of data and reads them separately.",
                           "The band is 10% of the length everywhere here, as in the benchmark; the window study is tier A of the roadmap.",
                           "The annealed winner and the lattice variants are read on TE and KL next to QE.",
                           "Every ingredient below has a mechanism and a control; the two temperatures (Sections 3 and 4) are the principled replacements.",
                           "Tear-and-rejoin came back as a rule by data support (GNG edges), a fracture criterion and a jigsaw step; the skeptics dropped all three (Section 3b). Its successor reads the walls off the U*-matrix after training.",
                           "Every map is judged with the Euclidean distance, with DTW and with its own; detection uses the own distance.",
                           "Every chart has a map-size axis; the strips compare a variant with its control at the same size and seed.",
                           "The growing DTW-SOM is the vehicle for the winners of this page; it is not rebuilt here."),
  check.names = FALSE)
tab(lessons, search = FALSE)
```

# 2. The two targets and the data of this page

**Partial discharges.** The project's PD set is airborne ultrasound of
corona, air-gap and void discharges plus ambient noise, 125 samples at
500 kS/s triggered by the current pulse, loudness over four orders of
magnitude, pressure used as asinh(p / 1 mPa). A PD engineer reads three
things from such a pulse: its *shape* (rise, ringing frequency, decay, a
reflection), its *loudness*, and, when a mains-phase reference exists, its
*phase* (the PRPD pattern). The map must therefore (a) use a distance that
reads shape without being ruled by loudness, (b) keep loudness somewhere it
can be read, as a channel or a per-unit statistic, and (c) answer two
questions: *is this a discharge?* (detection, with a false-alarm rate on
noise) and *which kind?* (classification).

**Movement.** A gesture or an activity comes from several axes at once, at
its own speed, held in a sensor that is never oriented the same way twice.
The map must align in time (DTW, soft-DTW, or a rigid shift), treat the axes
together (one alignment for all channels), and be read against the 1-NN DTW
baseline that any template matcher reaches.

`r if (all(sapply(datasets, function(d) d$source == "synthetic"))) "**This knit uses synthetic stand-ins**, because the project's data (som_bench/data/datasets.rds) is not in this checkout; put it there and the loader uses the real CBF, the real partial discharges and GunPoint instead (Section 4 shows the loader)." else "**This knit uses the project's own data** where the loader found it (see the sources below)."`
Each stand-in is built to have the difficulty that matters for its target:

```{r data-table}
tab(data.frame(`Data set` = sapply(datasets, `[[`, "title"), Source = sapply(datasets, `[[`, "source"),
               Items = sapply(datasets, function(d) nrow(d$X)), Length = sapply(datasets, `[[`, "L"), Channels = sapply(datasets, `[[`, "nch"),
               Classes = sapply(datasets, function(d) nlevels(d$cls)), `DTW band` = sapply(datasets, function(d) scales_of(d)$band),
               `1-NN accuracy, Euclidean` = f3(nn_floor["euclid", ]), `1-NN accuracy, DTW` = f3(nn_floor["dtw", ]),
               `What it is` = sapply(datasets, `[[`, "about"), check.names = FALSE), search = FALSE)
```

<p class="note">1-NN accuracy: leave-one-out nearest-neighbour classification, the floor for any
map that is then read with the same distance. The gap between the Euclidean and the DTW column says
how much a time alignment can buy on each set.</p>

## A few items of every class {.tabset .tabset-pills}

```{r examples, results = "asis"}
example_plot <- function(d, per_class = 4) {
  col <- class_colours(d$cls)
  show <- unlist(lapply(split(seq_along(d$cls), d$cls), function(i) head(i, per_class)))
  pl <- plot_ly()
  for (ch in seq_len(d$nch)) for (i in show) {
    y <- d$X[i, (ch - 1) * d$L + seq_len(d$L)]
    pl <- add_trace(pl, x = seq_len(d$L), y = y, type = "scatter", mode = "lines", xaxis = paste0("x", ch), yaxis = paste0("y", ch),
                    line = list(color = adjustcolor(col[[as.character(d$cls[i])]], 0.75), width = 1.3),
                    name = as.character(d$cls[i]), legendgroup = as.character(d$cls[i]), showlegend = ch == 1 && i == show[match(d$cls[i], d$cls[show])],
                    hovertemplate = paste0(as.character(d$cls[i]), ", item ", i, if (d$nch > 1) paste0(", axis ", ch), "<br>t = %{x}: %{y:.2f}<extra></extra>"))
  }
  lay <- list(margin = list(t = 20, b = 40), legend = list(orientation = "h", x = 0, y = -0.15), showlegend = TRUE)
  for (ch in seq_len(d$nch)) {
    dom <- c((ch - 1) / d$nch + 0.02 * (ch > 1), ch / d$nch - 0.02 * (ch < d$nch))
    lay[[paste0("xaxis", if (ch > 1) ch else "")]] <- ax(if (ch == d$nch) d$x_label else NULL, domain = dom, anchor = paste0("y", ch))
    lay[[paste0("yaxis", if (ch > 1) ch else "")]] <- ax(if (d$nch > 1) paste("axis", ch) else NULL, anchor = paste0("x", ch))
  }
  slim(do.call(chart, c(list(pl), lay)))
}
for (k in names(datasets)) {
  cat("\n\n### ", SHORT[[k]], "\n\n", sep = "")
  print(tagList(example_plot(datasets[[k]])))
  cat("\n\n")
}
```

# 3. The ideas

The brainstorm was run from seven angles, each idea then handed to a skeptic
from another angle who tried to refute it: is it already tested in the
project, is the mechanism real or a knob without a reason, are the
references right, is the control fair, does it plausibly matter for pulses
or for gestures. What survived is below; what was dropped is listed at the
end of the section with the reason. Tier A: build first; B: after A pays
off; C: a long shot worth one small experiment. Filter with the menus above
the table; the folded cells open on a click.

```{r ideas-table}
if (is.null(ideas)) {
  cat("The catalogue (next_steps/ideas/ideas.json) is not in this checkout.")
} else {
  fold <- function(s, n = 160) ifelse(nchar(s) > n, sprintf("<details><summary>%s&hellip;</summary>%s</details>", substr(s, 1, n), s), s)
  more <- function(i) sprintf(paste0("<details><summary>more</summary><p><b>For partial discharges.</b> %s</p><p><b>For movement.</b> %s</p>",
                                     "<p><b>How to test.</b> %s</p><p><b>Cost.</b> %s</p><p><b>Already tested?</b> %s</p><p><b>Risk.</b> %s</p>",
                                     "<p><b>The skeptic.</b> %s%s</p><p><b>References.</b> %s</p></details>"),
                              i$why_pd, i$why_movement, i$how_to_test, i$cost, i$already_tested, i$risk,
                              if (nzchar(i$problems)) i$problems else "no objection", if (nzchar(i$corrections)) paste0(" <i>Correction:</i> ", i$corrections) else "",
                              paste(unlist(i$anchors), collapse = "; "))
  # what this page already ran, by idea (the critic's point: a kept idea that the demonstrator tested is not new)
  RAN <- list(`L3-cosine-correlation-shape-loudness` = "Sections 6, 7: cosine and correlation (without the loudness coordinate)",
              `L6-sbd-one-global-shift` = "Sections 6, 7: SBD", `PD5-shift-not-warp` = "Sections 6, 7: SBD (the shift; not the narrow band)",
              `L1-manhattan-huber` = "Section 6: Manhattan, the clipped pull, and both", `L5-soft-dtw-gradient-update` = "Section 6: soft-DTW SOM",
              `L2-soft-dtw-free-energy` = "Section 6: soft-DTW SOM", `L1-free-energy-winner` = "Section 6: the annealed winner (soft arm only)",
              `L6-novelty-detection` = "Section 7, with non-map controls", `PD6-novelty-score` = "Section 7 (one threshold, no per-unit thresholds)",
              `L2-trust-continuity` = "Section 6: trustworthiness and continuity columns", `PD11-precision` = "Section 8, including raw pressure",
              `L8-multivariate-dtw-dependent-vs-independent` = "Section 6: dependent DTW only (the gestures)", `L3-magnification-control` = "Section 6: conscience (the equal-use end)")
  ran_of <- function(id) if (!is.null(RAN[[id]])) RAN[[id]] else "not yet"
  it <- do.call(rbind, lapply(ideas$ideas, function(i) data.frame(
    Tier = i$tier, Idea = sprintf("<b>%s</b><br><span class='note'>%s &middot; confidence %.2f &middot; references %s</span>%s", i$title, i$lens, i$confidence, i$anchor_status, more(i)),
    `In this page` = ran_of(i$id), `What it changes` = fold(paste(i$what_it_changes, i$mechanism)), `The control to beat` = fold(i$control),
    `Expected gain` = i$expected_gain, Lens = i$lens, Confidence = sprintf("%.2f", i$confidence), stringsAsFactors = FALSE, check.names = FALSE)))
  it <- it[order(match(it$Tier, c("A", "B", "C")), -as.numeric(it$Confidence)), ]
  chip_js <- JS("function(data, type) { if (type !== 'display') return data; return '<span class=\"chip ' + data + '\">' + data + '</span>'; }")
  filtered_table(it, c("Tier", "Lens", "In this page"), id = "ideas-table", escape = FALSE, dom = "ti", page_length = 100,
                 column_defs = list(list(render = chip_js, targets = 0), list(width = "24%", targets = 1), list(width = "9%", targets = 2),
                                    list(width = "26%", targets = 3), list(width = "18%", targets = 4), list(width = "10%", targets = 5),
                                    list(visible = FALSE, targets = c(6, 7))),
                 ordering = TRUE, paging = FALSE)
}
```

```{r ideas-headline, results = "asis"}
if (!is.null(ideas)) {
  cat("<div class='box'>\n\n**The synthesizer's headline**\n\n", ideas$headline, "\n\n</div>\n")
}
```

## 3a. Distances: what each one reads, and what its update does

A SOM has two places for a distance: the **contest** (which unit wins) and
the **update** (how the item pulls the prototype). The project learned the
hard way that the two must agree. For the pointwise distances the update
needs no change: the straight line from a prototype to an item is a
geodesic of every norm, so "move a share of the way towards the item" is
the right step for L1, L2, Chebyshev and Minkowski alike; what changes is
*which* prototype moves, and that is what matters.

* **Manhattan (L1)**: the winner is the prototype with the smallest *sum of
  absolute* differences. One loud sample (a spike, a glitch) counts in
  proportion to its size, not its square, so L1 is the first thing to try
  when outliers rule the Euclidean contest. Its natural partner is the
  **clipped pull** (the Huber step of Section 4): the pull of every sample is
  capped, so one loud item cannot drag a prototype.
* **Chebyshev (L-infinity)**: only the worst sample counts. It is the
  distance of a tolerance band and is sharp on sets where one feature decides
  (a step that is or is not there), blunt where everything is noisy.
* **Minkowski p**: between the two; p = 3 is the Euclidean contest with
  large differences weighed more.
* **Cosine and correlation**: the angle between item and prototype, with or
  without their means removed. Loudness drops out entirely. For partial
  discharges that is exactly the question *is it the same shape?* asked
  without *how loud?*, which is why Section 7 finds them where it does.
  Loudness then has to be kept elsewhere: as a per-unit statistic, or as an
  extra channel (the easiest test: append asinh(peak) as a 126th value with a
  weight).
* **Shape-based distance (SBD, from k-Shape)**: one global shift instead of
  a warp, found by cross-correlation; the update shifts the item by the best
  lag. It is the rigid cousin of DTW: it reads a pulse that came 3 samples
  late as the same pulse, costs L times less than DTW, and cannot stretch
  the inside of a series. For trigger-aligned pulses with a jitter it is the
  first thing to try before DTW.
* **DTW** (the project's) and **soft-DTW**: Section 3c.
* **Not built here, worth building**: dependent *vs* independent
  multivariate DTW for the gestures (one path for all axes against one path
  per axis), MSM and TWED (elastic measures with a cost for every edit, the
  strongest in the UCR bake-offs), the lower bounds that make DTW cheap at
  mapping time (the project has LB_Keogh), and distances on transformed
  series (the log-spectrum or the envelope of a pulse, where the ringing
  frequency and the decay become positions).

## 3b. The map: growing, tearing, rejoining, and what the skeptics left of it

The project's growing DTW-SOM won on fit, order and timing; its tearing map
lost to a smaller start radius because a link broken by *length* is the
same knob as a weaker pull, applied unevenly. Tear-and-rejoin came back
from three angles in the brainstorm (GNG edge ageing by data support, a
Griffith fracture criterion on the stored strain of a link, a jigsaw step
that re-seats torn pieces before they rejoin) and every version was dropped
by its skeptic: the ageing rule's lifetime is a second schedule with no
scale of its own, the fracture rule re-parametrises the length rule, and the
jigsaw rests on a hypothesis that the tear-and-join report contradicts
(rejoined late, the disorder stays). The honest successor leaves the pull
alone and finds the walls *after* training: cluster extraction from the
U*-matrix, distance times density (Ultsch), in place of the Ward cut, tier B.
The completeness critic reinstated one version at tier B, because the user
asked for it and the project's own untested list names cutting by data
support: GNG ageing on the hexagonal lattice with the lifetime set from the
data (an edge survives as long as its two units take, on average, to win
again), the pull following graph distance over the live edges, against the
fixed lattice, on the three blobs where tearing helped and on the 18 sets
where it did not, with the number of pieces counted as in the tearing lab.
That is the one run that closes the question either way. The critic also
reinstated the window per unit for the growing map (each unit's band set
from the shifts of the items it won, births inheriting the parent's band)
and kept the torus as the rim control for Heskes' winner and for the
topographic product, which reacts to the outline of a map.

What survived for the map itself:

* **Heskes' winner** (tier A): the unit whose whole neighbourhood fits the
  item best wins, the minimum over k of &Sigma;<sub>j</sub>
  h<sub>kj</sub> d<sub>j</sub><sup>2</sup>, instead of the nearest unit. It
  makes the online SOM a gradient descent on one energy (Heskes 1999), costs
  no extra alignment (the distances are there already), and is the pairing
  that could buy fit at a smaller start radius without the order cost the
  project measured. Control: the plain map at the same start radius and at
  the default.
* **A growth-stop rule and idle-unit pruning** in the growing map (tier B):
  stop when a new unit no longer lowers the Kaski-Lagus error on held-out
  seeds; remove a unit that wins nothing for a full pass.
* **LVQ fine-tuning** when labels exist (tier B): Kohonen's LVQ1 after the
  unsupervised map, prototypes moved a little towards their own class and
  away from the others. Control: the map read by Ward clustering; purity and
  ARI on held-out items.
* **Median and relational SOMs** (tier B): prototypes that are real series,
  so that any dissimilarity serves, including TWED and MSM, which have no
  barycentre to average with.
* **Online ordering, then batch convergence** (tier B): the online schedule
  down to its radius floor, then DBA passes at frozen winners: the fit of
  batch training without its disorder.
* **Dropped, and run here anyway** (Section 6): conscience (the chaotic-units
  report already ran the bounded control, and no unit was dead to rescue)
  and the torus (the rim errors it removes are the harmless kind). Neither
  beat its control in the demonstrator. Dropped and not run: DTW neural gas
  (between the SOM and k-means by construction), bagged maps (lattice
  alignment is ill-posed on the offset hexagonal grid), a learned band per
  unit (units with different bands do not compete on one scale).

## 3c. Physics: two temperatures, one tension, and the metaphors

```{r physics-bridge, results = "asis"}
if (!is.null(ideas) && nzchar(ideas$physics_bridge)) cat(ideas$physics_bridge, "\n\n")
```

The parts of that bridge that the demonstrator tests, in the words of the
code:

* <span class="chip exact">exact</span> **Soft-DTW is a free energy.** DTW
  takes the cheapest alignment; soft-DTW takes
  &minus;&gamma;&nbsp;log&nbsp;&Sigma;<sub>paths</sub>&nbsp;exp(&minus;cost/&gamma;),
  the free energy of all alignments at temperature &gamma; (Cuturi &
  Blondel 2017). Its gradient with respect to the prototype is the
  Boltzmann-weighted average of the alignments, the matrix E of the core,
  and the update of Section 4 averages the item with exactly those weights:
  the soft barycentre step. As &gamma; &rarr; 0 the path, the distance and the
  update all become DTW's (the check in Section 4 shows the two agree to
  10<sup>&minus;12</sup> at &gamma; = 0.001). One number, &gamma;, replaces the
  hard choice of one path with a smooth one; it costs about ten times DTW.
* <span class="chip exact">exact</span> **Deterministic annealing is a
  temperature on the winner.** Write the SOM's cost of assigning item x to
  unit k as E<sub>k</sub> = &Sigma;<sub>j</sub> h<sub>kj</sub> d(x, m<sub>j</sub>)<sup>2</sup>
  and let the item pull every unit with the Boltzmann weight
  P(k) &prop; exp(&minus;E<sub>k</sub>/T) (Graepel, Burger & Obermayer 1997;
  Heskes 1999 for why this is the winner that makes the SOM a gradient
  descent). At T &rarr; 0 the most probable unit takes all and the loop is
  Kohonen's; at high T every unit pulls alike and the map sits at the mean
  of the data, exactly the contracted state of the chaotic page's trap; as T
  falls the map unfolds, in order, through the phase transitions that split
  it along the principal axes of the data (Rose 1998). The radius schedule
  already plays part of this role, which is why the experiment, not the
  argument, decides whether annealing on top of a radius schedule buys
  anything: Section 6.
* <span class="chip exact">exact</span> **Heskes' winner is the zero-temperature
  limit of that assignment.** Take the unit with the lowest energy
  E<sub>k</sub> and let it win: the SOM then minimises one function, and the
  contest costs nothing beyond the distances already computed. The brainstorm
  ranks it tier A and the annealed version tier C: the hard arm first, the
  temperature only if the hard arm pays.
* <span class="chip exact">exact</span> **Magnification.** A 1-D SOM places
  prototypes with density p(x)<sup>2/3</sup> (Ritter & Schulten 1986), not
  p(x): it under-represents dense regions and over-represents sparse ones.
  Conscience pushes the exponent towards 1 (equal use of all units);
  Bauer, Der & Herrmann (1996) give a local learning rate that sets any
  exponent, including the information-optimal one for detection, where rare
  (sparse) pulses deserve more units, not fewer.
* <span class="chip metaphor">metaphor</span> **Springs and fracture.** The
  lattice as springs between prototypes with the neighbourhood as the
  stiffness, and a link that tears when its strain energy exceeds a Griffith
  threshold. Honest reading, and the skeptic's: the stiffness *is* the
  radius, and a tear by stored energy is a tear by length squared, the
  project's rule with another exponent. The elastic net of Durbin & Willshaw
  (1987) is the exact version of this energy for the ring; for a sheet it
  adds a knob, not a reason, and it was dropped.
* <span class="chip metaphor">metaphor</span> **Inertia (momentum).** A
  prototype that keeps a velocity. Cheap, so it ran in Section 6; the
  skeptic's reading is that momentum with a constant factor is a learning
  rate of &alpha;/(1 &minus; &mu;) with a lag, a reparametrisation of the
  schedule, and the results agree: no direction.
* <span class="chip metaphor">metaphor</span> **Replica exchange.** Several
  maps at different temperatures that swap; for a SOM this is *restarts with
  the best map kept*, which the ring experiments already found to be the
  working part of CHAOSOM. Dropped: the replicas would be deterministic-
  annealing runs, and the swap rule has nothing to decide that the energy
  of a single run does not. Keep the restarts, keep the best map by Heskes'
  energy, drop the exchange.

## 3d. Precision and the opposite direction

The user asked about fp64 data and an MPFR-precise map. Section 8 answers
with measurements; the reasoning is short. A SOM's state is the prototypes;
the only place a tiny numerical difference can grow is a **contest decided
by a near-tie**: if two prototypes are within the arithmetic's resolution of
each other, the winner can flip, and from that step on the two runs diverge
(the chaotic page saw the same with nudges of 10<sup>-3</sup>, and nothing
below 10<sup>-6</sup>). So the question is how often a contest is that
close, which the core now records at every step (the relative gap between
the first and the second distance). DTW adds a second place: the min over
three cells of a DP table, where a tie picks a path, but the cost is the
same whichever path is taken, and the warp differs only at ties that the
update then averages. The other direction is the one that matters for a
detector in the field: a map stored and evaluated in fp32, fp16 or
fixed-point integers on a microcontroller, with fp64 as the oracle. The
ladder of Section 8 is that oracle test run upwards. The brainstorm's one
tier-A idea here is about the data, not the arithmetic: asinh(p / 1 mPa) is
a companding law that turns four decades of loudness into about ten units,
and the question worth a table is how many bits of each class survive it
after the map's distance has done its work, which is also the question of
whether the map is a loudness meter that will not transfer out of the
anechoic chamber (Section 3a's loudness-only control answers the same
thing from the other side).

## 3e. Measuring across map sizes and for detection

* **Fair across sizes**: EV (a share, not a distance), the share of units
  used, purity against its random-units line, ARI and NMI (chance-corrected),
  and ranks within a size. Every fit measure that is a distance (QE, KL)
  falls with units; read it only within one size.
* **Trustworthiness and continuity** (Venna & Kaski 2001): are an item's k
  nearest neighbours *on the map* really its neighbours in the data
  (trustworthiness: no false neighbours), and do its k nearest neighbours
  *in the data* stay its neighbours on the map (continuity: no neighbour
  torn away)? Unlike TE they read the
  whole neighbourhood, not the runner-up alone, and unlike TP they do not
  react to the outline of the map. Computed here on the units' map
  distances against the Euclidean neighbourhoods of the items.
* **Detection**: train on the background, score every item by its distance
  to the winner, read the ROC. AUROC overall and per class: the per-class
  numbers show which discharges the map would miss.
* **Read the map on items it never saw.** Every ARI in the project's reports,
  and every one on this page, is on the training items with the number of
  classes given. The brainstorm's first experiment is to read held-out items
  (the seed's draw of items leaves a fold out), with a per-class confusion
  matrix and the detection AUROC, on the cached benchmarks: no retraining,
  and it may re-rank verdicts, because the project itself found part of
  DTW's gain in reading the map rather than training it.
* **One decision rule, stated once.** Wins seed by seed are counts, not
  p-values; under a two-sided sign test 7 of 10 seeds is p = 0.34 and 8 of
  10 is p = 0.11. So: a *win* is 9 or 10 of 10 seeds (or a paired Wilcoxon
  test at p < 0.05 whose confidence interval excludes zero); 7 or 8 of 10 is
  *undecided* and triggers a 30-seed rerun; *level* is an equivalence
  verdict, not an absence of evidence. With the `r p$seeds` seeds of this
  knit only `r p$seeds` of `r p$seeds` (p = `r signif(2 / 2^p$seeds, 2)`) is a
  statement at all, and a difference inside the seed-to-seed spread is noise
  whatever its sign. The project's summary page would lose some of its
  "helps" verdicts under this rule; that is the first thing to recount.
* **Consensus over seeds** is the cheapest ARI the cached maps can give: a
  co-association matrix over the ten Ward cuts of ten seeds, cut once, is an
  ensemble that needs no training; its controls are the single map with the
  lowest KL and a consensus of ten k-means runs, so that the gain is
  attributed to the map or to the averaging. Trustworthiness and continuity as computed here need at
  least 19 sampled items for their normalisation; the sample is 300.

```{r ideas-dropped, results = "asis"}
if (!is.null(ideas) && length(ideas$dropped)) {
  cat("<details><summary>Ideas the skeptics dropped, and why</summary>\n\n")
  dr <- do.call(rbind, lapply(ideas$dropped, function(d) data.frame(Idea = d$title, Lens = d$lens, `Why it was dropped` = d$why_dropped, check.names = FALSE, stringsAsFactors = FALSE)))
  print(tab(dr, search = TRUE))
  cat("\n\n</details>\n\n")
}
if (!is.null(ideas) && length(ideas$gaps)) {
  cat("<details><summary>What the completeness critic still found missing</summary>\n\n")
  for (g in ideas$gaps) cat(sprintf("* **%s.** %s *Fix:* %s\n", g$title, g$why_it_matters, g$suggested_fix))
  cat("\n\n</details>\n\n")
}
```

# 4. The demonstrator

One online SOM, written once in C++ as a template over the scalar type, with
the distance as an argument of both the contest and the update, and six
ingredients (the clipped pull, conscience, the annealed winner, momentum,
the rectangular sheet, the torus) that each switch one idea on. It is not the project's core and
does not try to match it to the last bit; it follows the same conventions
(hexagonal lattice, Gaussian neighbourhood, learning rate 0.5 &rarr; 0.01,
kohonen's start radius &rarr; 0 used as 0.5 below 1, the start and the
order of items drawn from the seed), so that a variant and its control see
the same items in the same order from the same start.

```{r variants-table}
vt <- do.call(rbind, lapply(VARIANTS, function(v) data.frame(
  Variant = v$label, Group = v$group, Distance = v$mode, Lattice = paste(v$topo, if (v$torus) "torus" else "sheet"),
  Control = VARIANTS[[v$family]]$label, `What it changes` = v$what, check.names = FALSE, stringsAsFactors = FALSE)))
vt$Control[vt$Variant == vt$Control] <- ""
tab(vt, search = FALSE)
```

Settings that come from the data so that one value means the same thing
everywhere: the DTW band and the SBD shift are 10% of the length; soft-DTW's
&gamma; is 0.1 of the variance of one sample; the clipped pull caps each
component at one standard deviation of a sample; the annealing temperature
starts at the mean squared distance of the items to their mean (under the
map's own distance) and falls geometrically by a factor 1000 over the first
90% of the steps, then the winner is hard; conscience uses &beta; = 0.25 and
a win-share memory of 10<sup>4</sup> steps; momentum keeps half of the last
step.

**Checks.** The core's DTW equals a plain R implementation (squared local
cost, Sakoe-Chiba band) on 200 random pairs; soft-DTW converges to DTW
squared as &gamma; &rarr; 0 and its expected-alignment warp to DTW's path
average (10<sup>-12</sup> at &gamma; = 0.001); the pointwise distances equal
R's `dist()`; dependent multichannel DTW scales as it should; the online SOM
in fp64 equals the same loop written in R to the last bit; the topographic
product is 0 on an ordered chain and negative on a scrambled one; the
Kaski-Lagus way along an ordered chain is the sum of its links. The script
is `next_steps/tests/test_core.R`.

The code, in full:

```{Rcpp core-code, eval = FALSE, code = xfun::read_utf8("next_steps/ns_core.cpp")}
```

```{r data-code, eval = FALSE, code = xfun::read_utf8("next_steps/R/ns_data.R")}
```

```{r maps-code, eval = FALSE, code = xfun::read_utf8("next_steps/R/ns_maps.R")}
```

```{r quality-code, eval = FALSE, code = xfun::read_utf8("next_steps/R/ns_quality.R")}
```

```{r bench-code, eval = FALSE, code = xfun::read_utf8("next_steps/R/ns_bench.R")}
```

```{r precision-code, eval = FALSE, code = xfun::read_utf8("next_steps/R/ns_precision.R")}
```

```{r report-code, eval = FALSE, code = xfun::read_utf8("next_steps/R/ns_report.R")}
```

# 5. The maps, at four sizes

One figure per data set, seed 1: pick a variant in the menu and see its map
at `r paste(sprintf('%d &times; %d', SIZES, SIZES), collapse = ', ')` units side by
side. Each cell is tinted with the class of most of the items it wins (white:
none), and carries its prototype as a small curve (the first axis for the
gestures). Hover a cell for the items it won by class. Within one size,
every variant starts from the same items and sees them in the same order;
across sizes the draws differ, because a bigger map draws more start items.

## Maps {.tabset .tabset-pills}

```{r gallery, results = "asis"}
gallery_plot <- function(k) {
  d <- datasets[[k]]; col <- class_colours(d$cls); W <- 940; H <- 270; ns <- length(SIZES)
  vids <- names(VARIANTS); pl <- plot_ly(width = W, height = H); vis_group <- c()
  lay <- list(margin = list(t = 36, b = 10, l = 10, r = 10), showlegend = FALSE)
  for (si in seq_along(SIZES)) {
    s <- SIZES[si]; dom <- c((si - 1) / ns + 0.01, si / ns - 0.01)
    lay[[paste0("xaxis", if (si > 1) si else "")]] <- list(domain = dom, anchor = paste0("y", si), visible = FALSE, range = c(0.2, s + 1.3), fixedrange = TRUE)
    lay[[paste0("yaxis", if (si > 1) si else "")]] <- list(anchor = paste0("x", si), visible = FALSE, range = c(0.2, s * 0.87 + 1), fixedrange = TRUE, scaleanchor = paste0("x", si))
  }
  for (v in vids) for (si in seq_along(SIZES)) {
    s <- SIZES[si]; f <- fits[[paste(k, v, s, sep = "|")]]
    if (is.null(f)) next
    grid <- make_grid(s, s, VARIANTS[[v]]$topo, VARIANTS[[v]]$torus)
    fill <- cell_fill(f$bmu, d$cls, grid$K, col)
    counts <- table(factor(f$bmu, levels = seq_len(grid$K)), d$cls)
    hover <- sapply(seq_len(grid$K), function(u) sprintf("<b>%s, %d x %d</b><br>unit %d: %d items<br>%s", VARIANTS[[v]]$label, s, s, u, sum(counts[u, ]),
                                                          paste(sprintf("%s %d", colnames(counts), counts[u, ]), collapse = ", ")))
    pitch <- (W / ns - 20) / (s + 1.1)                                   # pixels per lattice unit in this panel
    cp <- cell_paths(grid)
    pl <- add_trace(pl, x = round(cp$x, 3), y = round(cp$y, 3), type = "scatter", mode = "lines", xaxis = paste0("x", si), yaxis = paste0("y", si),
                    line = list(color = "#c3c2b7", width = 0.6), hoverinfo = "skip", visible = v == vids[1])
    pl <- add_trace(pl, x = grid$pos[, 1], y = grid$pos[, 2], type = "scatter", mode = "markers", xaxis = paste0("x", si), yaxis = paste0("y", si),
                    marker = list(symbol = if (grid$topo == "hex") "hexagon" else "square", size = if (grid$topo == "hex") 0.72 * pitch else 0.9 * pitch,
                                  color = fill, line = list(width = 0)),
                    text = hover, hoverinfo = "text", visible = v == vids[1])
    cv <- proto_curves(f$M, grid, d$L, d$nch, channel = 1, y_range = range(d$X[, seq_len(d$L)]))
    pl <- add_trace(pl, x = cv$x, y = cv$y, type = "scatter", mode = "lines", xaxis = paste0("x", si), yaxis = paste0("y", si),
                    line = list(color = "#D1343B", width = 1), hoverinfo = "skip", visible = v == vids[1])
    vis_group <- c(vis_group, v, v, v)
  }
  buttons <- lapply(vids, function(v) list(method = "restyle", label = VARIANTS[[v]]$label, args = list(list(visible = vis_group == v))))
  for (si in seq_along(SIZES)) lay$annotations[[si]] <- list(xref = "paper", yref = "paper", x = (si - 0.5) / ns, y = 1.0, xanchor = "center", yanchor = "bottom",
                                                           text = sprintf("%d x %d", SIZES[si], SIZES[si]), showarrow = FALSE, font = list(size = 13, color = INK2))
  lay$updatemenus <- list(menu(buttons, type = "dropdown", x = 0, y = 1.08))
  slim(do.call(chart, c(list(pl), lay)))
}
for (k in names(datasets)) {
  cat("\n\n### ", SHORT[[k]], "\n\n", sep = "")
  print(tagList(gallery_plot(k)))
  cat("\n\n")
}
```

<p class="note">Class colours as in the examples of Section 2. The soft-DTW
map stops at 10 &times; 10 (its 14 &times; 14 map would cost about ten
minutes per seed on this machine).</p>

# 6. Every measure, every size

## 6a. One measure across the sizes

Pick the measure (the menu) and the data set (the buttons). Each line is a
variant, each point the mean over `r p$seeds` seeds at one map size, with the
range over the seeds as a band on hover. Measures that fall with the number
of units (QE, KL) are shown within one data set only; read them at one size
at a time.

```{r size-chart}
size_chart <- function() {
  vids <- names(VARIANTS); keys <- names(datasets)
  series <- list()                                            # one trace per data set and variant
  for (k in keys) for (v in vids) series[[paste(k, v)]] <- list(k = k, v = v)
  yvals <- lapply(VIEWS, function(vw) lapply(series, function(sv) {
    y <- sapply(SIZES, function(s) { x <- val(sv$k, sv$v, s, vw$m, vw$j); if (length(x)) mean(x, na.rm = TRUE) else NA })
    I(y) }))
  texts <- lapply(VIEWS, function(vw) lapply(series, function(sv) I(sapply(SIZES, function(s) {
    x <- val(sv$k, sv$v, s, vw$m, vw$j)
    if (!length(x) || all(is.na(x))) return("")
    sprintf("<b>%s</b>, %d x %d (%d units)<br>%s: %s (seeds %s to %s)", VARIANTS[[sv$v]]$label, s, s, s^2, vw$label,
            formatC(mean(x, na.rm = TRUE), digits = 3, format = "fg", flag = "#"), formatC(min(x, na.rm = TRUE), digits = 3, format = "fg", flag = "#"),
            formatC(max(x, na.rm = TRUE), digits = 3, format = "fg", flag = "#")) }))))
  pl <- plot_ly(height = 640)
  v0 <- names(VIEWS)[1]
  for (nm in names(series)) {
    sv <- series[[nm]]; i <- match(sv$v, vids)
    pl <- add_trace(pl, x = SIZES^2, y = yvals[[v0]][[nm]], text = texts[[v0]][[nm]], type = "scatter", mode = "lines+markers", name = VARIANTS[[sv$v]]$label,
                    legendgroup = sv$v, showlegend = sv$k == keys[1], visible = sv$k == keys[1], hovertemplate = "%{text}<extra></extra>",
                    line = list(color = PAL[i], width = if (VARIANTS[[sv$v]]$group == "control") 3.2 else 1.8, dash = if (VARIANTS[[sv$v]]$group == "control") "solid" else if (VARIANTS[[sv$v]]$group == "distance") "solid" else "dot"),
                    marker = list(color = PAL[i], size = 7, symbol = if (VARIANTS[[sv$v]]$mode %in% c("dtw", "softdtw")) "square" else "circle"))
  }
  measure_buttons <- lapply(names(VIEWS), function(id) list(method = "update", label = VIEWS[[id]]$label,
    args = list(list(y = unname(yvals[[id]]), text = unname(texts[[id]])),
                list(yaxis.title.text = VIEWS[[id]]$label, yaxis.type = if (VIEWS[[id]]$kind == "rel") "log" else "linear", yaxis.autorange = TRUE))))
  data_buttons <- lapply(keys, function(k) list(method = "restyle", label = SHORT[[k]], args = list(list(visible = sapply(series, function(sv) sv$k == k)))))
  chart(pl, xaxis = ax("units on the map (log scale)", type = "log", tickvals = SIZES^2, ticktext = sprintf("%d x %d", SIZES, SIZES)),
        yaxis = ax(VIEWS[[v0]]$label), legend = list(orientation = "v", x = 1.02, y = 1, yanchor = "top", font = list(size = 11), tracegroupgap = 1),
        updatemenus = list(menu(measure_buttons, type = "dropdown", x = 0, y = 1.02), menu(data_buttons, x = 0.55, y = 1.02)),
        margin = list(t = 60, r = 10, b = 50)) |> slim()
}
size_chart()
```

<p class="note">Thick lines: the two controls (online SOM, DTW-SOM). Solid
thin lines: distances; dotted: ingredients and lattices. Squares: the maps
trained with DTW or soft-DTW. Legend entries toggle a variant on all data
sets.</p>

## 6b. Every variant against its control, by data set and size

Blue: the variant beat its family's control (same seed, same start, same
order) on more seeds than it lost; red: the other way round. Pick the measure.

```{r heatmap, fig.height = 6.2}
vids <- setdiff(names(VARIANTS), c("euclid", "dtw"))
cols <- expand.grid(size = SIZES, k = names(datasets), stringsAsFactors = FALSE)
col_lab <- sprintf("%s %d x %d", SHORT[cols$k], cols$size, cols$size)
cells <- lapply(names(VIEWS), function(id) {
  Z <- matrix(NA_real_, length(vids), nrow(cols)); H <- matrix("", length(vids), nrow(cols))
  for (i in seq_along(vids)) for (j in seq_len(nrow(cols))) {
    a <- tryCatch(adv_of(cols$k[j], vids[i], cols$size[j], id), error = function(e) NULL)
    if (is.null(a) || !nrow(a) || all(is.na(a$adv))) next
    Z[i, j] <- sum(a$adv > 0, na.rm = TRUE) - sum(a$adv < 0, na.rm = TRUE)
    H[i, j] <- sprintf("<b>%s</b> on %s<br>%s: better than %s on %d, worse on %d of %d seeds<br>mean %s against %s", VARIANTS[[vids[i]]]$label, col_lab[j],
                       VIEWS[[id]]$label, VARIANTS[[VARIANTS[[vids[i]]]$family]]$label, sum(a$adv > 0, na.rm = TRUE), sum(a$adv < 0, na.rm = TRUE), sum(!is.na(a$adv)),
                       formatC(mean(a$v, na.rm = TRUE), digits = 3, format = "fg", flag = "#"), formatC(mean(a$c, na.rm = TRUE), digits = 3, format = "fg", flag = "#"))
  }
  list(Z = Z, H = H)
})
names(cells) <- names(VIEWS)
ylab <- sapply(vids, function(v) VARIANTS[[v]]$label)
div_scale <- list(c(0, "#e34948"), c(0.25, "#f3aeac"), c(0.5, "#f0efec"), c(0.75, "#9ec5f4"), c(1, "#2a78d6"))
hm_buttons <- lapply(names(VIEWS), function(id) list(method = "restyle", label = VIEWS[[id]]$label, args = list(list(z = list(cells[[id]]$Z), text = list(cells[[id]]$H)))))
plot_ly(x = col_lab, y = ylab, z = cells[[1]]$Z, text = cells[[1]]$H, type = "heatmap", hoverinfo = "text", colorscale = div_scale,
        zmin = -p$seeds, zmax = p$seeds, xgap = 2, ygap = 2,
        colorbar = list(title = list(text = "seeds better minus worse than the control", side = "top", font = list(size = 12, color = INK2)),
                        orientation = "h", x = 0.5, xanchor = "center", y = -0.32, yanchor = "top", len = 0.5, thickness = 12, outlinewidth = 0,
                        tickvals = c(-p$seeds, 0, p$seeds), ticktext = c("worse on all seeds", "tie", "better on all seeds"), tickfont = list(color = INK2))) |>
  chart(xaxis = ax(NULL, tickangle = -35, showgrid = FALSE), yaxis = ax(NULL, autorange = "reversed", showgrid = FALSE, ticksuffix = "  "),
        shapes = lapply(length(SIZES) * seq_len(length(datasets) - 1), function(at) list(type = "line", xref = "x", yref = "paper", x0 = at - 0.5, x1 = at - 0.5, y0 = 0, y1 = 1, line = list(color = INK2, width = 1.2))),
        updatemenus = list(menu(hm_buttons, type = "dropdown", x = 0, y = 1.02)), margin = list(t = 60, r = 20, b = 150, l = 200)) |> slim()
```

<p class="note">The torus is compared with the rectangular sheet, the
rectangular sheet and all other Euclidean variants with the online SOM, the
DTW variants with the DTW-SOM. An empty cell was not run (soft-DTW at
14 &times; 14).</p>

## 6c. Seed by seed, at `r REF` &times; `r REF`

Each dot is one seed: how much better the variant did than its control with
the same start and the same order of items. Right of the line: the variant
was better. Pick the measure and the data set.

```{r strip}
strip_chart <- function(size = REF) {
  vids <- setdiff(names(VARIANTS), c("euclid", "dtw")); keys <- names(datasets); nr <- length(vids); ypos <- nr - seq_along(vids)
  vlab <- sapply(vids, v_lab)
  build <- function(k, id) {
    D <- do.call(rbind, lapply(seq_along(vids), function(r) { a <- tryCatch(adv_of(k, vids[r], size, id), error = function(e) NULL); if (is.null(a) || !nrow(a)) NULL else cbind(row = r, a) }))
    if (is.null(D)) return(NULL)
    vw <- VIEWS[[id]]; ratio <- vw$kind == "rel"
    D$x <- if (ratio) D$c / D$v else D$adv
    if (ratio && !vw$higher) D$x <- D$c / D$v else if (ratio) D$x <- D$v / D$c
    M <- aggregate(x ~ row, D, function(v) if (ratio) exp(mean(log(v))) else mean(v))
    set.seed(3)
    list(x = I(D$x), y = I(ypos[D$row] + runif(nrow(D), -0.22, 0.22)), col = I(ifelse(D$adv > 0, "#eb6834", ifelse(D$adv < 0, "#2a78d6", AXIS))),
         text = I(sprintf("<b>%s</b>, seed %d<br>variant %s, control %s", vlab[D$row], D$seed, formatC(D$v, digits = 3, format = "fg", flag = "#"), formatC(D$c, digits = 3, format = "fg", flag = "#"))),
         mx = I(M$x), my = I(ypos[M$row]), mtext = I(sprintf("<b>%s</b>: %s %.3g", vlab[M$row], if (ratio) "geometric mean" else "mean", M$x)),
         type = if (ratio) "log" else "linear", ref = if (ratio) 1 else 0,
         axis = if (ratio) paste(vw$label, ": control / variant (log scale; right of 1 = variant better)") else paste(vw$label, ": variant minus control (right of 0 = variant better)"))
  }
  B <- lapply(keys, function(k) { r <- lapply(names(VIEWS), build, k = k); names(r) <- names(VIEWS); r })
  names(B) <- keys
  ref_line <- function(x0) list(list(type = "line", xref = "x", yref = "paper", x0 = x0, x1 = x0, y0 = 0, y1 = 1, line = list(color = INK2, width = 2)))
  pl <- plot_ly(height = 160 + 40 * nr)
  for (k in keys) {
    b <- B[[k]][[1]]
    pl <- add_trace(pl, x = b$x, y = b$y, type = "scatter", mode = "markers", text = b$text, hovertemplate = "%{text}<extra></extra>", visible = k == keys[1],
                    marker = list(color = b$col, size = 10, opacity = 0.85, line = list(color = "#ffffff", width = 1.5)))
    pl <- add_trace(pl, x = b$mx, y = b$my, type = "scatter", mode = "markers", text = b$mtext, hovertemplate = "%{text}<extra></extra>", visible = k == keys[1],
                    marker = list(color = INK, size = 14, symbol = "line-ns", line = list(color = INK, width = 3)))
  }
  measure_buttons <- lapply(names(VIEWS), function(id) {
    xs <- list(); ys <- list(); ts <- list(); cs <- list()
    for (k in keys) { b <- B[[k]][[id]]; xs <- c(xs, list(b$x, b$mx)); ys <- c(ys, list(b$y, b$my)); ts <- c(ts, list(b$text, b$mtext)); cs <- c(cs, list(b$col, INK)) }
    b1 <- B[[keys[1]]][[id]]
    list(method = "update", label = VIEWS[[id]]$label, args = list(list(x = xs, y = ys, text = ts, marker.color = cs),
                                                                   list(xaxis.title.text = b1$axis, xaxis.type = b1$type, xaxis.autorange = TRUE, shapes = ref_line(b1$ref))))
  })
  data_buttons <- lapply(keys, function(k) list(method = "restyle", label = SHORT[[k]], args = list(list(visible = rep(keys == k, each = 2)))))
  chart(pl, xaxis = ax(B[[keys[1]]][[1]]$axis), yaxis = ax(NULL, tickvals = ypos, ticktext = unname(vlab), range = c(-0.6, nr - 0.4), ticksuffix = "  ", zeroline = FALSE),
        shapes = ref_line(B[[keys[1]]][[1]]$ref), showlegend = FALSE,
        annotations = list(list(xref = "paper", yref = "paper", x = 0, y = 1.01, xanchor = "left", yanchor = "bottom", showarrow = FALSE, text = "<span style='color:#2a78d6'>&#9679;</span> control better", font = list(size = 13, color = INK)),
                           list(xref = "paper", yref = "paper", x = 1, y = 1.01, xanchor = "right", yanchor = "bottom", showarrow = FALSE, text = "variant better <span style='color:#eb6834'>&#9679;</span>", font = list(size = 13, color = INK))),
        updatemenus = list(menu(measure_buttons, type = "dropdown", x = 0, y = 1.06), menu(data_buttons, x = 0.6, y = 1.06)), margin = list(t = 80, b = 60, l = 200)) |> slim()
}
strip_chart(REF)
```

## 6d. All numbers

Means over the seeds, every variant, every data set, every size, every judge.
Filter with the menus above the table to one size and one judge before
comparing QE or KL; click a heading to sort.

```{r all-numbers}
agg <- aggregate(cbind(QE, EV, TE, KL, TP, used, purity, ARI, NMI, trust, cont, seconds, comparisons, not_nearest, gap_median, gap_tiny) ~ dataset + variant + size + judge,
                 data = bench, FUN = function(v) mean(v, na.rm = TRUE), na.action = na.pass)
agg$`Data set` <- factor(SHORT[agg$dataset], levels = unname(SHORT)); agg$Variant <- factor(sapply(agg$variant, v_lab), levels = sapply(VARIANTS, `[[`, "label"))
agg$Map <- factor(sprintf("%d x %d", agg$size, agg$size), levels = sprintf("%d x %d", SIZES, SIZES)); agg$Judge <- factor(JUDGE_LAB[agg$judge], levels = JUDGE_LAB)
out <- agg[order(agg$`Data set`, agg$Map, agg$Variant, agg$Judge), c("Data set", "Map", "Variant", "Judge", "QE", "EV", "TE", "KL", "TP", "used", "purity", "ARI", "NMI", "trust", "cont", "seconds", "not_nearest", "gap_tiny")]
names(out)[names(out) == "used"] <- "Units used"; names(out)[names(out) == "purity"] <- "Purity"; names(out)[names(out) == "trust"] <- "Trust"; names(out)[names(out) == "cont"] <- "Continuity"
names(out)[names(out) == "seconds"] <- "Seconds"; names(out)[names(out) == "not_nearest"] <- "Winner not nearest"; names(out)[names(out) == "gap_tiny"] <- "Contests closer than 1e-6"
out$`Data set` <- as.character(out$`Data set`); out$Map <- as.character(out$Map); out$Variant <- as.character(out$Variant); out$Judge <- as.character(out$Judge)
filtered_table(out, c("Data set", "Map", "Variant", "Judge"), id = "all-numbers", page_length = 17,
               signif_cols = c("QE", "KL", "TP", "Seconds"),
               round_cols = c("EV", "TE", "Units used", "Purity", "ARI", "NMI", "Trust", "Continuity", "Winner not nearest", "Contests closer than 1e-6"))
```

```{r results-text, include = FALSE}
# the sentences of the box below, from the results
best_by <- function(k, view, size = REF) { b <- best_variant(k, size, view); sprintf("%s (%s)", b$label, formatC(b$value, digits = 2, format = "f")) }
ctrl_of <- function(k, view, size = REF) formatC(mean_of(k, "euclid", size, view), digits = 2, format = "f")
soft_vs_dtw <- sapply(names(datasets), function(k) wins_of(k, "softdtw", REF, "qe_dtw"))
anneal_te <- sapply(names(datasets), function(k) wins_of(k, "anneal", REF, "te_own")); anneal_qe <- sapply(names(datasets), function(k) wins_of(k, "anneal", REF, "qe_eu"))
cos_pd <- wins_of("pd", "cosine", REF, "ari_own"); sbd_pd <- wins_of("pd", "sbd", REF, "ari_own"); l1_pd <- wins_of("pd", "manhattan", REF, "ari_own")
dtw_pd_wins <- wins_vs("pd", "dtw", "euclid", REF, "ari_own")
sbd_ge <- wins_of("gestures", "sbd", REF, "ari_own"); dtw_ge <- wins_vs("gestures", "dtw", "euclid", REF, "ari_own"); cor_ge <- wins_of("gestures", "correlation", REF, "ari_own")
hub_ari <- sapply(names(datasets), function(k) wins_of(k, "huber", REF, "ari_own")); l1h_ari <- sapply(names(datasets), function(k) wins_of(k, "l1_huber", REF, "ari_own"))
ann_ari <- sapply(names(datasets), function(k) wins_of(k, "anneal", REF, "ari_own")); dan_ari <- sapply(names(datasets), function(k) wins_of(k, "dtw_anneal", REF, "ari_own"))
tor_qe <- sapply(names(datasets), function(k) wins_of(k, "torus", REF, "qe_eu")); tor_ari <- sapply(names(datasets), function(k) wins_of(k, "torus", REF, "ari_own"))
rect_ari <- sapply(names(datasets), function(k) wins_of(k, "rect", REF, "ari_own")); con_ari <- sapply(names(datasets), function(k) wins_of(k, "conscience", REF, "ari_own"))
mom_ari <- sapply(names(datasets), function(k) wins_of(k, "momentum", REF, "ari_own"))
l1_qe <- sapply(names(datasets), function(k) wins_of(k, "manhattan", REF, "qe_eu")); l1_ari <- sapply(names(datasets), function(k) wins_of(k, "manhattan", REF, "ari_own"))
hub_qe <- sapply(names(datasets), function(k) wins_of(k, "huber", REF, "qe_eu")); con_used <- sapply(names(datasets), function(k) wins_of(k, "conscience", REF, "used_own"))
tor_te <- sapply(names(datasets), function(k) wins_of(k, "torus", REF, "te_own")); rect_te <- sapply(names(datasets), function(k) wins_of(k, "rect", REF, "te_own"))
mom_qe <- sapply(names(datasets), function(k) wins_of(k, "momentum", REF, "qe_eu")); mom_te <- sapply(names(datasets), function(k) wins_of(k, "momentum", REF, "te_own"))
soft_te <- sapply(names(datasets), function(k) wins_of(k, "softdtw", REF, "te_dtw")); soft_ari <- sapply(names(datasets), function(k) wins_of(k, "softdtw", REF, "ari_dtw"))
dan_te <- sapply(names(datasets), function(k) wins_of(k, "dtw_anneal", REF, "te_own")); dan_qe <- sapply(names(datasets), function(k) wins_of(k, "dtw_anneal", REF, "qe_dtw"))
fmt_w <- function(w) if (is.null(dim(w))) sprintf("%d of %d", w[["better"]], w[["n"]]) else paste(sprintf("%d of %d", w["better", ], w["n", ]), collapse = ", ")
secs_ratio <- function(v, k = "pd", size = REF) mean_of(k, v, size, "secs") / mean_of(k, "euclid", size, "secs")
```

<div class="box">

Counts are seeds on which the variant beat its control, out of `r p$seeds`,
at `r REF` &times; `r REF`, in the order `r paste(SHORT, collapse = ", ")`.

* **Distances.** The classes of the partial-discharge stand-in were found
  best by `r best_by("pd", "ari_own")`, of the gestures by
  `r best_by("gestures", "ari_own")`, of CBF by `r best_by("cbf", "ari_own")`
  (ARI with each map's own distance; online SOM `r ctrl_of("pd", "ari_own")`,
  `r ctrl_of("gestures", "ari_own")` and `r ctrl_of("cbf", "ari_own")`).
  Against the online SOM on the PD classes: cosine `r fmt_w(cos_pd)`, SBD
  `r fmt_w(sbd_pd)`, Manhattan `r fmt_w(l1_pd)`, the DTW-SOM `r fmt_w(dtw_pd_wins)`;
  on the gestures: SBD `r fmt_w(sbd_ge)`, the DTW-SOM `r fmt_w(dtw_ge)`,
  correlation `r fmt_w(cor_ge)`. SBD costs
  `r sprintf("%.0f", secs_ratio("sbd"))` times the online SOM, DTW
  `r sprintf("%.0f", secs_ratio("dtw"))` times, soft-DTW
  `r sprintf("%.0f", secs_ratio("softdtw"))` times.
* **Manhattan and the clipped pull.** Manhattan against the online SOM:
  classes `r fmt_w(l1_ari)`; fit judged with the Euclidean distance
  `r fmt_w(l1_qe)` (its rival's own distance, where it cannot win by
  construction). The clipped pull alone: fit `r fmt_w(hub_qe)`, classes
  `r fmt_w(hub_ari)`; Manhattan with the clipped pull: classes
  `r fmt_w(l1h_ari)`. On data with few outliers an L1 contest is a different
  contest, not a better one; the clipped pull is the cheaper of the two to
  keep.
* **The two temperatures.** Soft-DTW against the DTW-SOM, both read with
  DTW: fit `r fmt_w(soft_vs_dtw)`, order (TE) `r fmt_w(soft_te)`, classes
  (ARI) `r fmt_w(soft_ari)`. The annealed winner on the Euclidean map: order
  `r fmt_w(anneal_te)`, fit `r fmt_w(anneal_qe)`, classes `r fmt_w(ann_ari)`;
  on the DTW-SOM: order `r fmt_w(dan_te)`, fit `r fmt_w(dan_qe)`, classes
  `r fmt_w(dan_ari)`. The annealed contest chose a unit other than the nearest
  in `r pct(mean(bench$not_nearest[bench$variant == "anneal" & bench$judge == "own"]))`
  of the steps: it is a different map, and on these data at this size not a
  better one. Both are worth one more test where a wrong early winner costs
  most, inside the growing map.
* **Lattices.** The torus against the rectangular sheet: order (TE)
  `r fmt_w(tor_te)`, fit `r fmt_w(tor_qe)`, classes `r fmt_w(tor_ari)`; the
  rectangular sheet against the hexagonal one: order `r fmt_w(rect_te)`,
  classes `r fmt_w(rect_ari)`. With four neighbours instead of six, TE counts
  more runner-ups as errors, so compare the two sheets on KL and on trust,
  not on TE alone.
* **Conscience** used more of the map on `r fmt_w(con_used)` seeds and found
  the classes better on `r fmt_w(con_ari)`; **momentum**: fit (Euclidean QE)
  `r fmt_w(mom_qe)`, order `r fmt_w(mom_te)`, classes `r fmt_w(mom_ari)`.

</div>

# 7. Detection: a map trained on noise

For partial discharges the first question is not *which kind* but *is it
one*. The simplest detector a SOM offers: train on the background (the noise
class only), then score every waveform by its distance to its winner; a
discharge should land far from every prototype of noise. Each variant,
7 &times; 7 maps, `r p$seeds` seeds, the AUROC of that score for discharges
against noise, overall and per kind:

```{r detection, fig.height = 5}
BASE_LAB <- c(base_peak = "Control: loudness (peak of the series)", base_1nn_euclid = "Control: nearest noise item, Euclidean",
              base_1nn_cosine = "Control: nearest noise item, cosine", base_1nn_dtw = "Control: nearest noise item, DTW")
det_lab <- function(v) if (v %in% names(VARIANTS)) VARIANTS[[v]]$label else BASE_LAB[[v]]
dv <- names(VARIANTS); da <- aggregate(. ~ variant, data = det[, c("variant", grep("^AUROC", names(det), value = TRUE))], FUN = mean)
da <- da[order(-da$AUROC), ]
pl <- plot_ly(height = 480)
for (v in da$variant) {
  d <- det[det$variant == v, ]; is_base <- !(v %in% dv)
  pl <- add_trace(pl, x = d$AUROC, y = rep(det_lab(v), nrow(d)), type = "scatter", mode = "markers", name = det_lab(v), showlegend = FALSE,
                  marker = list(color = if (is_base) MUTED else PAL[match(v, dv)], size = if (is_base) 12 else 10, symbol = if (is_base) "diamond" else "circle",
                                opacity = 0.85, line = list(color = "#ffffff", width = 1.5)),
                  text = sapply(seq_len(nrow(d)), function(i) sprintf("<b>%s</b>%s<br>AUROC %.3f<br>%s", det_lab(v), if (is_base) "" else sprintf(", seed %d", d$seed[i]), d$AUROC[i],
                                 paste(sprintf("%s %.2f", sub("AUROC_", "", grep("^AUROC_", names(d), value = TRUE)), as.numeric(d[i, grep("^AUROC_", names(d))])), collapse = ", "))),
                  hovertemplate = "%{text}<extra></extra>")
}
chart(pl, xaxis = ax("AUROC of the score as a discharge alarm (1 = perfect, 0.5 = chance)", range = c(0, 1.02)),
      yaxis = ax(NULL, categoryorder = "array", categoryarray = rev(sapply(da$variant, det_lab)), ticksuffix = "  "),
      shapes = list(list(type = "line", x0 = 0.5, x1 = 0.5, yref = "paper", y0 = 0, y1 = 1, line = list(color = AXIS, dash = "dot"))), margin = list(t = 20, l = 250)) |> slim()
dat <- da; dat$variant <- sapply(dat$variant, det_lab); names(dat) <- c("Score", "AUROC, all discharges", paste("AUROC,", sub("AUROC_", "", names(dat)[-(1:2)])))
tab(dat, search = FALSE, digits = setNames(as.list(rep(3, ncol(dat) - 1)), names(dat)[-1]))
det_som <- det[det$variant %in% dv, ]
det_mean <- function(v) mean(det$AUROC[det$variant == v])
det_best <- names(which.max(sapply(dv, det_mean)))
base_best <- names(which.max(sapply(names(BASE_LAB), det_mean)))
```

Grey diamonds are the controls that are not maps: the loudness of the
series as the alarm (the first detector a PD engineer would try), and the
distance to the nearest noise item under the Euclidean, cosine and DTW
distances (a noise item is scored against the other noise items). A map
earns its place only above those.

<div class="box">

* The `r v_lab(det_best)` map is the best alarm among the maps (AUROC
  `r f3(det_mean(det_best))`); the online SOM reaches `r f3(det_mean("euclid"))`.
  The best control that is not a map is `r tolower(sub("Control: ", "", BASE_LAB[[base_best]]))`
  (`r f3(det_mean(base_best))`); the loudness threshold alone scores
  `r f3(det_mean("base_peak"))`. So
  `r if (det_mean(det_best) > det_mean(base_best) + 0.01) "the map adds something over the nearest-neighbour score under the same distance" else if (det_mean(det_best) >= det_mean(base_best) - 0.01) "the map and the nearest-neighbour score under the same distance are level, so the distance is the finding, not the map" else "the nearest-neighbour score beats the map, so the distance is the finding and the map costs accuracy for its summary"`. Per kind, the Euclidean
  alarm scores `r f3(det_corona_eu)` on corona`r if (det_corona_eu < 0.5) ": below 0.5 means a corona pulse lies *closer* to the noise prototypes than noise does, because it is quiet and a Euclidean distance is mostly loudness" else ""`.
  An amplitude-blind distance (cosine, correlation, SBD) asks about the shape
  instead`r if (det_corona_eu < 0.5) ", and the quiet pulses come back" else ""`.
* This is the one place on this page where the distances differ by a wide
  margin, and the one that matters first in the field. The roadmap's first
  experiment is this test on the real waveforms, with a false-alarm rate at
  a fixed detection rate next to the AUROC.

</div>

# 8. The precision ladder

```{r ladder-table}
lt <- lad_tab
lt$data <- c(cbf = sprintf("%s, %d x %d, %d passes", small_cbf$title, LADDER$cbf$size, LADDER$cbf$size, LADDER$cbf$rlen),
             pd = sprintf("partial-discharge stand-in (600 x 125, asinh scale), %d x %d, %d passes", LADDER$pd$size, LADDER$pd$size, LADDER$pd$rlen),
             pd_raw = sprintf("the same pulses as raw pressure in mPa (no asinh), %d x %d, %d passes", LADDER$pd_raw$size, LADDER$pd_raw$size, LADDER$pd_raw$rlen))[lt$data]
lt$drift <- formatC(lt$drift, digits = 2, format = "e"); lt$flip_share <- NULL
names(lt) <- c("Data", "Distance", "Arithmetic", "Reference", "First step with another winner", "Steps with another winner", "Largest difference of a prototype", "QE", "TE", "Seconds")
tab(lt, search = FALSE, digits = list(QE = 4, TE = 4, Seconds = 2))
```

The same seed, the same start, the same order of items, trained at every
precision; "another winner" is counted against the `r p$mpfr_bits`-bit MPFR
run, which repeats the core's operations one by one in the MPFR library
(`next_steps/ns_mpfr.cpp`); the row "Rmpfr check" is the same loop written
in R with the Rmpfr package, an independent check of that code. The largest
difference of a prototype is over all units and all values at the end.

```{r ladder-curves, fig.height = 4.6}
cv <- do.call(rbind, lapply(names(LD), function(k) cbind(data = k, ladder[[k]]$curves)))
cv$drift[cv$drift <= 0] <- NA
precs <- unique(cv$precision)
pl <- plot_ly(height = 400)
for (dk in names(LD)) for (mo in c("euclid", "dtw")) for (pr in precs) {
  d <- cv[cv$data == dk & cv$mode == mo & cv$precision == pr, ]
  if (!nrow(d) || all(is.na(d$drift))) next
  pl <- add_trace(pl, x = d$step, y = d$drift, type = "scatter", mode = "lines+markers", name = pr, legendgroup = pr, showlegend = dk == "cbf" && mo == "euclid",
                  visible = dk == "cbf" && mo == "euclid", line = list(color = PAL[match(pr, precs)], width = 2), marker = list(color = PAL[match(pr, precs)], size = 6),
                  text = sprintf("%s, %s<br>step %d: largest difference %.2e", pr, mo, d$step, d$drift), hovertemplate = "%{text}<extra></extra>")
}
groups <- unlist(lapply(names(LD), function(dk) lapply(c("euclid", "dtw"), function(mo) sapply(precs, function(pr) { d <- cv[cv$data == dk & cv$mode == mo & cv$precision == pr, ]; nrow(d) > 0 && !all(is.na(d$drift)) }))), recursive = FALSE)
keys4 <- as.vector(outer(c("euclid", "dtw"), names(LD), function(m, d) paste(d, m)))[c(1, 2, 3, 4, 5, 6)]
labs4 <- as.vector(t(outer(unname(LD), c("Euclidean", "DTW"), paste, sep = ", ")))
vis_all <- unlist(lapply(seq_along(groups), function(i) rep(i, sum(groups[[i]]))))
buttons <- lapply(seq_along(keys4), function(i) list(method = "restyle", label = labs4[i], args = list(list(visible = vis_all == i))))
chart(pl, xaxis = ax("training step (log scale)", type = "log"), yaxis = ax("largest difference from the reference (log)", type = "log", exponentformat = "power"),
      updatemenus = list(menu(buttons, x = 0, y = 1.03)), legend = list(orientation = "h", x = 0, y = -0.25), margin = list(t = 50, b = 70)) |> slim()
```

```{r gap-hist, fig.height = 3.8}
gh <- do.call(rbind, lapply(names(LD), function(k) rbind(data.frame(data = LD[[k]], mode = "Euclidean", gap = ladder[[k]]$gap), data.frame(data = LD[[k]], mode = "DTW", gap = ladder[[k]]$gap_dtw))))
gh <- gh[is.finite(gh$gap) & gh$gap > 0, ]
pl <- plot_ly(height = 330)
i <- 0
for (dk in unique(gh$data)) for (mo in unique(gh$mode)) {
  i <- i + 1; g <- gh$gap[gh$data == dk & gh$mode == mo]
  pl <- add_trace(pl, x = log10(g), type = "histogram", name = paste(dk, mo), opacity = 0.55, nbinsx = 60, marker = list(color = PAL[i]),
                  hovertemplate = paste(dk, mo, "<br>log10 gap %{x}: %{y} steps<extra></extra>"))
}
chart(pl, barmode = "overlay", xaxis = ax("log10 of the relative gap between the first and the second distance in a contest"), yaxis = ax("steps"),
      shapes = list(list(type = "line", x0 = log10(2^-23), x1 = log10(2^-23), yref = "paper", y0 = 0, y1 = 1, line = list(color = INK2, dash = "dot")),
                    list(type = "line", x0 = log10(2^-52), x1 = log10(2^-52), yref = "paper", y0 = 0, y1 = 1, line = list(color = INK2, dash = "dot"))),
      annotations = list(list(x = log10(2^-23), y = 1, yref = "paper", text = "fp32 resolution", showarrow = FALSE, xanchor = "left", font = list(size = 11, color = INK2)),
                         list(x = log10(2^-52), y = 1, yref = "paper", text = "fp64 resolution", showarrow = FALSE, xanchor = "left", font = list(size = 11, color = INK2))),
      legend = list(orientation = "h", x = 0, y = -0.3), margin = list(t = 20, b = 70)) |> slim()
```

```{r ladder-text, include = FALSE}
row_of <- function(dk, mo, pr) lad_tab[lad_tab$data == dk & lad_tab$mode == mo & lad_tab$precision == pr, ]
fp32c <- row_of("cbf", "euclid", "fp32"); fp32p <- row_of("pd", "euclid", "fp32"); fp32d <- row_of("pd", "dtw", "fp32")
fp64c <- row_of("cbf", "euclid", "fp64"); fp64p <- row_of("pd", "euclid", "fp64"); fp64d <- row_of("pd", "dtw", "fp64"); fp64cd <- row_of("cbf", "dtw", "fp64")
kah <- row_of("pd", "euclid", "fp64 + Kahan"); d32 <- row_of("pd", "euclid", "fp32 data, fp64 arithmetic"); d32c <- row_of("cbf", "euclid", "fp32 data, fp64 arithmetic")
q_te <- function(dk, mo) range(lad_tab$TE[lad_tab$data == dk & lad_tab$mode == mo]); q_qe <- function(dk, mo) range(lad_tab$QE[lad_tab$data == dk & lad_tab$mode == mo])
gap_pd <- gap_small[, "pd"]
```

<div class="box">

* **fp64, 80-bit, 128-bit and MPFR agree.** On the small CBF the fp64
  Euclidean map chose `r flip_words(fp64c$flips, ladder$cbf$S)` as the
  `r p$mpfr_bits`-bit run, and its prototypes end within `r e1(fp64c$drift)` of
  the reference; on the PD stand-in `r flip_words(fp64p$flips, ladder$pd$S)`
  (drift `r e1(fp64p$drift)`); under DTW, `r fp64cd$flips` and `r fp64d$flips`
  steps differ. The Rmpfr loop agreed with the MPFR core
  `r if (nrow(rmpfr_row) && rmpfr_row$flips[1] == 0 && rmpfr_row$drift[1] == 0) "on every winner and every bit of every prototype" else sprintf("except for %d winners and a drift of %s", rmpfr_row$flips[1], e1(rmpfr_row$drift[1]))`
  (`r sprintf("%.0f", rmpfr_row$seconds[1])` s against `r sprintf("%.1f", mpfr_secs)` s in C). Kahan summation changed
  `r if (kah$flips == fp64p$flips && abs(kah$drift - fp64p$drift) < 1e-14) "nothing" else sprintf("the count to %d and the drift to %s", kah$flips, e1(kah$drift))`.
  The measures of all these runs agree to the digits shown (TE from
  `r f3(q_te("pd", "euclid")[1])` to `r f3(q_te("pd", "euclid")[2])` on the PD stand-in across
  every arithmetic including fp32).
* **fp32 drifts, and it does not matter.** The fp32 Euclidean run
  `r first_words(fp32c, ladder$cbf$S)` on the small CBF, and
  `r first_words(fp32p, ladder$pd$S)` on the PD stand-in
  (under DTW: `r if (is.na(fp32d$first_flip)) "no other winner either" else sprintf("another winner first at step %d", fp32d$first_flip)`).
  The curves show the drift growing from 10<sup>-6</sup> as the training goes
  on. Where a winner does flip, the map that follows is a different map, not
  a worse one: the spread between seeds is far larger than the spread between
  arithmetics.
* **Raw pressure is the case built to fail, and it is the one to watch.**
  Without the asinh scale the loudest sample of an air-gap pulse squared is
  10<sup>8</sup> times a corona sample squared, so one squared distance mixes
  numbers that fp32 cannot add. On the same pulses as raw pressure fp32
  `r if (is.na(raw32$first_flip)) sprintf("still chose the reference's winner at every step (drift %s)", e1(raw32$drift)) else sprintf("chose another winner at %d of %s steps, the first at step %d, and drifted to %s", raw32$flips, format(ladder$pd_raw$S, big.mark = ","), raw32$first_flip, e1(raw32$drift))`;
  under DTW `r if (is.na(raw32d$first_flip)) "no other winner" else sprintf("another winner first at step %d", raw32d$first_flip)`;
  fp64 `r if (raw64$flips == 0) "still agreed with the reference at every step" else sprintf("differed at %d steps", raw64$flips)`.
  The lesson is the project's asinh scale, not a precision upgrade: compand
  the data and fp32 is enough; feed a map raw pressure over four decades and
  fp32 is not, but neither is the map, because it is then a loudness meter.
* **fp64 data make no difference either.** Rounding every value of the data
  (items and start alike) to fp32 and training in fp64 gave `r d32$flips`
  other winners on the PD stand-in and `r d32c$flips` on CBF: the data of a
  16-bit or 24-bit converter carry nothing that fp64 arithmetic cannot keep.
* **Why.** `r if (gap_pd["fp32"] == 0) sprintf("No contest of the PD run came closer than fp32's resolution (2<sup>-23</sup>); the closest had a relative gap of %s", e1(gap_pd["min"])) else sprintf("A contest closer than fp32's resolution (2<sup>-23</sup>) happened in %s of the steps of the PD run", pct(gap_pd["fp32"]))`,
  and `r if (gap_pd["fp64"] == 0) "none came near fp64's (2<sup>-52</sup>)" else sprintf("%s came closer than fp64's (2<sup>-52</sup>)", pct(gap_pd["fp64"]))`;
  the histogram puts nearly every contest between 10<sup>-3</sup> and 1.
  Winners flip at near-ties, near-ties are rare, and when one flips the map
  that follows is one more seed, not a broken one.
* **So:** fp64 is the right precision to train in; MPFR is the oracle to
  prove it (`r sprintf("%.0f", mpfr_secs)` s for `r format(ladder$cbf$S, big.mark = ",")`
  steps of a 7 &times; 7 map, against milliseconds), useful once per core
  change, not for training. The experiment to run next is downwards: the
  trained map stored as fp16 or as 16-bit fixed point and evaluated on a
  microcontroller, with the fp64 winners as the truth and the count of
  flips as the score, because that is where a PD detector will live.

</div>

# 9. The roadmap

```{r tiers, results = "asis"}
if (!is.null(ideas) && length(ideas$tiers)) {
  title_of <- function(id) { i <- Filter(function(x) x$id == id, ideas$ideas); if (length(i)) i[[1]]$title else id }
  for (t in ideas$tiers) {
    cat(sprintf("**Tier %s.** %s\n\n", t$tier, gsub("\n+", "\n\n", t$rationale)))
    cat(sprintf("<p class='note'>Ideas of this tier (%d): %s.</p>\n\n", length(t$ids), paste(sapply(t$ids, title_of), collapse = " &middot; ")))
  }
  cat("<details><summary>The five to start with for partial discharges, and the five for movement</summary>\n\n")
  pick <- function(ids) paste(sapply(ids, function(id) { i <- Filter(function(x) x$id == id, ideas$ideas); if (length(i)) i[[1]]$title else id }), collapse = "; ")
  cat("* **Partial discharges:** ", pick(ideas$pd_top5), "\n* **Movement:** ", pick(ideas$movement_top5), "\n\n</details>\n\n", sep = "")
}
```

<div class="box">

**How to run each experiment**, in the project's way: the same seeds, the
variant against the control that does the same job without the new
ingredient, the start and the order of items drawn from the seed, every map
judged with the Euclidean distance, with DTW and with its own, QE next to TE
(or KL), ARI when the classes are known and now on held-out items, AUROC and
the per-class misses when the question is detection; counts of seeds better
and worse with a sign test, never a mean alone; and the same experiment at
two map sizes, because Section 6 shows the ranking move with the size.

1. **Reading and statistics first** (tier A, a day, no training): the
   cached benchmarks of the project and of this page read on held-out items,
   with per-class confusion, the novelty AUROC against the non-map controls
   of Section 7, consensus over the seeds, and the decision rule of Section
   3e. The real PD set has 850 untouched waveforms per class beyond the 150
   the benchmark drew: train on the 150, test on the 850, and report the
   false-alarm rate at 1% with real counts. Expect some of the summary's
   "helps" verdicts to become "undecided", and a 30-seed rerun of the few
   that matter.
2. **The representation ladder for partial discharges** (tier A, two days):
   on the real waveforms, a map on pulse-shape features (rise, decay,
   dominant frequency, energy, the time-frequency map of Contin and
   colleagues), one on the envelope and the log-spectrum, one on the series
   with loudness as a separate weighted coordinate, and the loudness-only
   control (log peak, Ward cut), against the Euclidean and the DTW-SOM. If a
   five-number map matches the DTW-SOM's ARI, the elastic family is moot on
   this data. The demonstrator's evidence for this order: the amplitude-blind
   distances led detection by a wide margin (Section 7). One question to the
   data's owners first: do the recordings carry the mains phase, the apparent
   charge or a timestamp per pulse? If they do, the phase-resolved pattern is
   the field's standard feature and belongs in the ladder as a circular
   coordinate, which is where the torus earns its place.
3. **Shift, not warp** (tier A, a day): SBD and a 2 to 3% DTW band against
   the 10% DTW-SOM on the real pulses, with a per-class histogram of the
   best shift. Trigger-aligned pulses jitter by a few samples; that is a
   shift, not a warp. In the demonstrator SBD was level with or ahead of the
   DTW-SOM on the PD classes at a fraction of the cost (Section 6).
4. **Heskes' winner** (tier A, two days): the energy-consistent winner at
   half and a quarter of the start radius, against the plain map at the same
   radius and at the default, on QE and TE together. The only change to the
   map's own rule that the skeptics kept at tier A.
5. **The input layer for movement** (tier A, two days): dependent against
   independent multichannel DTW on real gesture sets, with the check that
   dependent DTW at band 0 equals the stacked Euclidean map to the last bit,
   the 1-NN DTW ceiling, per-class held-out confusion with leave-subjects-out
   folds, and the engineer's control that PD has and movement lacked: a
   Euclidean map on windowed features (per-axis mean, spread, energy,
   dominant frequency, inter-axis correlations, the magnitude channel), with
   a supervised ceiling on the same features.
6. **Tier B, after the tables above**: soft-DTW's expected-alignment update
   inside the growing map (hard winner kept), priced warps (TWED, MSM),
   slope-constrained steps and open ends for gestures (with the project's
   derivative-DTW variant as the control row, since it attacked the same
   pinch and won once, on the solar I-V curves), LVQ fine-tuning, U*-matrix
   cluster extraction, a growth-stop rule and idle-unit pruning, the window
   per unit, GNG edges with a data-set lifetime (the one tear-and-rejoin
   left standing), median and relational maps.
   **One rule for all of them:** any ingredient that wins on the fixed map is
   re-run inside the growing DTW-SOM, the project's best map, with its births
   using the ingredient's own distance or barycentre, against the plain
   growing map with the same seeds.
7. **Tier C, one small run each**: 1-D Wasserstein on energy profiles,
   coarse-to-fine training on the envelope, a streaming growing map, the
   fixed-point Manhattan map for a field detector, the annealed winner, and
   the precision ladder as a falsification test on the real data.
8. **Not worth the time** on the evidence here and in the project: MPFR
   training, momentum, tearing by any rule tried so far, chaos of any kind,
   the torus, conscience, a PCA start.

</div>

# 10. Limits

* **Synthetic stand-ins.** The partial discharges and the gestures of this
  knit are generated to have the difficulty that matters (loudness over four
  decades with shape as the signal; speed, amplitude and orientation changes
  with shape as the signal), not to be the real data. Every number above is
  a statement about the demonstrator on these sets; the roadmap's first step
  repeats the study on the real waveforms. The loader swaps them in when
  `som_bench/data/datasets.rds` is present.
* **One setting of everything**: 10% band, `r p$rlen` passes, one schedule,
  &gamma; = 0.1 of a sample's variance, one temperature schedule, one clip,
  one conscience, `r p$seeds` seeds. Each of these would move some results;
  the project's window study shows how much one of them can.
* **The demonstrator is a fixed-size online map.** The growing DTW-SOM, the
  project's best, is where the winners of this page should go; it is not
  rebuilt here, and nothing above says how an ingredient behaves while a map
  grows.
* **The idea catalogue is a brainstorm**, verified by skeptics for mechanism,
  prior testing and fair controls, with references marked where they could
  not be checked. It ranks by expected value per hour of your time, which is
  a judgement, not a measurement.
* **A contest's gap** (Section 8 and the last column of the big table) is
  the second distance minus the first, over the magnitude of the first; for
  soft-DTW, whose values can be negative, it is a ratio of the same kind but
  not on the same scale as a distance.
* **The measure list and the size grid are this page's, not the project's.**
  The timing error (CBF), the spread (colours), the pieces (torn maps) and
  the tour gap (rings) are not computed here, and the sizes are 5, 7, 10 and
  14 a side where the benchmark used one rule of about 5&radic;N units. The
  roadmap's tables should carry every set-specific measure whenever its set
  is run, and keep CBF and the colours as sanity sets, because they catch
  smearing and shape loss that the PD and gesture sets do not.
* **Times** are from `r cores` parallel processes on this machine
  (`r format(n_maps, big.mark = ",")` maps in `r sprintf("%.0f", cpu_min)` CPU
  minutes); the number of comparisons is the fair cost.

# References

* Ahalt, S. C., Krishnamurthy, A. K., Chen, P. & Melton, D. E. (1990). Competitive learning algorithms for vector quantization. *Neural Networks* 3(3), 277-290.
* Bauer, H.-U. & Pawelzik, K. R. (1992). Quantifying the neighborhood preservation of self-organizing feature maps. *IEEE Transactions on Neural Networks* 3(4), 570-579.
* Bauer, H.-U., Der, R. & Herrmann, M. (1996). Controlling the magnification factor of self-organizing feature maps. *Neural Computation* 8(4), 757-771.
* Cuturi, M. & Blondel, M. (2017). Soft-DTW: a differentiable loss function for time-series. *Proceedings of the 34th International Conference on Machine Learning (ICML)*, 894-903.
* DeSieno, D. (1988). Adding a conscience to competitive learning. *IEEE International Conference on Neural Networks*, vol. 1, 117-124.
* Durbin, R. & Willshaw, D. (1987). An analogue approach to the travelling salesman problem using an elastic net method. *Nature* 326, 689-691.
* Fritzke, B. (1995). A growing neural gas network learns topologies. *Advances in Neural Information Processing Systems 7*, 625-632.
* Graepel, T., Burger, M. & Obermayer, K. (1997). Phase transitions in stochastic self-organizing maps. *Physical Review E* 56(4), 3876-3890.
* Heskes, T. (1999). Energy functions for self-organizing maps. In Oja, E. & Kaski, S. (eds), *Kohonen Maps*, 303-315. Elsevier.
* Hubert, L. & Arabie, P. (1985). Comparing partitions. *Journal of Classification* 2, 193-218.
* Kaski, S. & Lagus, K. (1996). Comparing self-organizing maps. *ICANN 96*, Lecture Notes in Computer Science 1112, 809-814.
* Kiviluoto, K. (1996). Topology preservation in self-organizing maps. *IEEE International Conference on Neural Networks (ICNN 1996)*, 294-299.
* Kohonen, T. (2013). Essentials of the self-organizing map. *Neural Networks* 37, 52-65.
* Paparrizos, J. & Gravano, L. (2015). k-Shape: efficient and accurate clustering of time series. *Proceedings of the 2015 ACM SIGMOD International Conference on Management of Data*, 1855-1870.
* Ritter, H. & Schulten, K. (1986). On the stationary state of Kohonen's self-organizing sensory mapping. *Biological Cybernetics* 54, 99-106.
* Rose, K. (1998). Deterministic annealing for clustering, compression, classification, regression, and related optimization problems. *Proceedings of the IEEE* 86(11), 2210-2239.
* Saito, N. (1994). *Local feature extraction and its applications using a library of bases*. PhD thesis, Yale University.
* Sakoe, H. & Chiba, S. (1978). Dynamic programming algorithm optimization for spoken word recognition. *IEEE Transactions on Acoustics, Speech, and Signal Processing* 26(1), 43-49.
* Somervuo, P. & Kohonen, T. (1999). Self-organizing maps and learning vector quantization for feature sequences. *Neural Processing Letters* 10(2), 151-159.
* Venna, J. & Kaski, S. (2001). Neighborhood preservation in nonlinear projection methods: an experimental study. *ICANN 2001*, Lecture Notes in Computer Science 2130, 485-491.
* Vesanto, J. & Alhoniemi, E. (2000). Clustering of the self-organizing map. *IEEE Transactions on Neural Networks* 11(3), 586-600.
